---
title: "ISA 444: Business Forecasting"
subtitle: "17 - Stationarity"
author: Fadel M. Megahed
date: 'Spring 2022'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    latex_engine: xelatex
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.8,
                      fig.height= 2.9,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, tidyquant, xtable, magrittr, fpp2, ggpubr, tseries)
library(ggplot2); theme_set(theme_bw(base_size = 8))
```

# Preface


### What we Covered Prior to Spring Break {.allowframebreaks}

\begin{block}{\textbf{Main Learning Outcomes}}
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define the population mean, and variance of a random variable.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define the population covariance, and correlation between two random variables.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define the population autocovariance and autocorrelation of a random variable.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Use sample estimates of the population mean, variance, covariance, and correlation.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Explain the properties of the large sample distribution of the sample ACF.} \\
	  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Use the large sample distribution of the sample ACF to identify significant autocorrelation in a time series.}
\end{block}


\begin{block}{\textbf{Main Learning Outcomes}}
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Determine if a sample ACF plot ÔÇťcuts offÔÇŁ or ÔÇťdies downÔÇŁ.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Explain how sample partial autocorrelation is calculated.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Determine if a sample PACF plot ÔÇťcuts offÔÇŁ or ÔÇťdies downÔÇŁ.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Use the sample ACF plot to identify a nonstationary time series.}
\end{block}


### Recap: HW 11 - ACF and PACF

Per your feedback post exam 01, let us discuss the solutions to [HW 10](https://miamioh.instructure.com/courses/167327/quizzes/449730?module_item_id=3678615) over the next few minutes.

**As usual**, please do not hesitate to ask me questions.


### Recall: A Formal Definition for Stationarity

**Weak Stationarity:** A weakly stationary time series is a finite variance process such that:  

  1. The mean, $\mu_t$, is constant and does not depend on the time $t$; and   
  2. The autocovariance function, $\gamma(s, t)$ depends on $s$ and $t$ only through their difference $|s-t|$.  
  
We will use the term ÔÇťstationaryÔÇŁ to refer to weak stationarity.  

  - The concept of weak stationarity forms the basis of much of the foundation for time series modeling.  
  - The fundamental properties (1 & 2) required for weak stationarity are satisfied by many of the models that are widely used.


### Recall: A Visual Explanation of Stationarity

```{r Stationary, echo=FALSE}
ar.sim = arima.sim(model=list(ar=c(.6)),n=500)

p1 = ar.sim %>% autoplot() + theme_bw() + labs(y = 'AR(1): 0.6')
p2 = acf(ar.sim, plot = FALSE, lag.max = 12) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```



### Recall: A Visual Explanation of Nonstationarity

```{r nonStationary, echo=FALSE}
google = tq_get("GOOG", from = Sys.Date() - 731, to = Sys.Date() -1) %>% select(date, adjusted)
google %>% ggplot(aes(x= date, y = adjusted)) + geom_line() + theme_bw() -> g1

acfGoogle = acf(google$adjusted, plot = F)

acfGoogle %>% autoplot() + theme_bw() -> g2

pacf(google$adjusted, plot = F) -> pacfGoogle

pacfGoogle %>% autoplot() + theme_bw() -> g3

ggarrange(g1, g2, ncol = 1) 
```


### Learning Objectives for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Apply transformations to a nonstationary time series to bring it into stationarity.}
			\item \textbf{Recognize and explain a random walk model (both with and without a drift).}
			\item \textbf{Recognize a random walk model from an ACF plot.}
			\item \textbf{Conduct formal tests for stationarity using the ADF and KPSS tests.}
	\end{itemize}
\end{block}




# Stationarity

### What to do when we have a Nonstationary series?

In order to model a time series, it must usually be in a state of stationarity. If the time series is not stationary, you must transform it to achieve stationarity.\footnotemark

Successive **differencing** is typically used to achieve stationarity.

**First Differences:** $y_t^\prime = y_t - y_{t-1}$.

**Second Differences"** $y_t'' = y_t^\prime - y_{t-1}^\prime$.

\footnotetext{Slides are based on \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's Handouts} for ISA 444, Spring 2020.}


### A Live Example: Examining the US GNP

```{r gnp, echo=FALSE}
gnp = tq_get('GNP', get = 'economic.data')

gnp$date %<>% ymd() 
gnp %<>% mutate(diff1 = c(NA, diff(price)),
                diff2 = c(NA, NA,  diff(price, differences = 2)))

df =  pivot_longer(data= gnp, cols = c(3,4,5)) 

df %>% ggplot(aes(x= date, y = value, group = name, color = name)) + 
  geom_line(size = 1) + facet_wrap(~ name, ncol = 1, scales = "free_y") + theme_bw() +
  theme(legend.position = 'none') + labs(x = 'Date') 
```


### So Why Does Differences Work?

*Because many nonstationary time series have features of a random walk.*



### Random Walk Model [1]

**Random Walk with Drift:** A random walk is a model such that successive differences (first differences) are independent. 

The classic random walk model: $$Y_t = Y_{t-1} + \epsilon_t$$

A random walk with a drift: $$Y_t = \delta + Y_{t-1} + \epsilon_t$$

**Notes:** When $\delta = 0$, the value of the current observation is just the value of the prior observation plus random noise.


### Random Walk Model [2]

```{r rw1, echo = FALSE}
set.seed(1)
randomWalk = data.frame(obsNum = 1:500, rawData = rnorm(500)) %>% 
  mutate(walk = cumsum(rawData), walkWdrift = cumsum(rawData + 0.2))

df = randomWalk %>% select(obsNum, walk, walkWdrift) %>% pivot_longer(c(2,3), names_to = 'TimeSeries')

df %>% ggplot(aes(x = obsNum, y = value, group = TimeSeries, color = TimeSeries)) +
  geom_line() + geom_smooth(method = 'lm') + theme_bw() + theme(legend.position = 'bottom')
```



### Random Walk Model [3]

```{r rw2, echo = FALSE}
acf(randomWalk$walk, plot = FALSE) %>% autoplot() +
  theme_bw() + labs(title = 'ACF of the Random Walk (no drift)')
```


### Random Walk Model [4]

```{r rw3, echo = FALSE}
acf(randomWalk$walkWdrift, plot = FALSE) %>% autoplot() +
  theme_bw() + labs(title = 'ACF of the Random Walk (With drift)')
```



# Formal Tests for Stationarity

### Basic Idea

**Unit Root Test:** One way to objectively determine if differencing is required is to use a unit root test. A unit root is a feature of a stochastic process that indicates a time series is nonstationary.


### Augmented Dickey Fuller (ADF) Test [1]

The **Augmented Dickey Fuller (ADF) Test** tests whether or not there is a unit root. The hypotheses are as follows:  

  $\qquad \qquad$ Ho: The series is nonstationary  
  
  $\qquad \qquad$ Ha: The series is stationary  
  
Thus, if we have a *SMALL p-value*, we reject the null hypothesis and conclude **STATIONARITY**.



### Augmented Dickey Fuller (ADF) Test [2]

In R, we will use the function `adf.test()` from the package [tseries](https://cran.r-project.org/web/packages/tseries/tseries.pdf). 

```{r adfTest}
pacman::p_load(tseries)
adf.test(gnp$price)
```

\textcolor{miamired}{\textbf{So what do we conclude from the test above?}}


### Kwiatkowski-Phillips-Schmidt-Shin (KPSS) Test [1]

The **Kwiatkowski-Phillips-Schmidt-Shin (KPSS) Test** tests whether or not there is a unit root. The hypotheses are as follows:  

  $\qquad \qquad$ Ho: The series is stationary  

  $\qquad \qquad$ Ha: The series is nonstationary  

Thus, if we have a *SMALL p-value*, we reject the null hypothesis and conclude **NONSTATIONARITY**.



### Kwiatkowski-Phillips-Schmidt-Shin (KPSS) Test [2]

In R, we will use the function `kpss.test()` from the package [tseries](https://cran.r-project.org/web/packages/tseries/tseries.pdf). 

```{r kpssTest}
pacman::p_load(tseries)
kpss.test(gnp$price)
```

\textcolor{miamired}{\textbf{So what do we conclude from the test above?}}



### Successive Kwiatkowski-Phillips-Schmidt-Shin (KPSS) Tests

As a followup to the case when the `kpss.test()` is rejected (or alternatively when you do not reject the `adf.test()`), you can utilize the `ndiffs()` from the package [forecast](https://cran.r-project.org/web/packages/forecast/forecast.pdf), which uses a series of the KPSS tests in a sequence to determine the appropriate number of first differences required for a nonseasonal time series. 

`ndiffs()` returns the number of first differences needed to achieve stationarity according to the KPSS test.


```{r ndiffs}
pacman::p_load(fpp2) # fpps loads the forecast package as well
ndiffs(gnp$price)
```

According to the `ndiffs() function`, **two successive differences are recommended to transform the GNP data into stationarity**.



# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Apply transformations to a nonstationary time series to bring it into stationarity.}
			\item \textbf{Recognize and explain a random walk model (both with and without a drift).}
			\item \textbf{Recognize a random walk model from an ACF plot.}
			\item \textbf{Conduct formal tests for stationarity using the ADF and KPSS tests.}
	\end{itemize}
\end{block}



### Things to Do to Prepare for the Exam Next Monday

 - Thoroughly read Chapters 1-4 of our reference book.
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered. 
 
 - Go over the interactive tutorials 



---

\maketitle