---
title: "26 - Class Notes"
author: "Fadel M Megahed"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    paged_df: true
    theme: simplex
    code_folding: show
    code_download: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives for Today's Class {-}

  1. Use a simple linear regression model for trend adjustment.
  2. Interpret regression diagnostic plots.  
  3. Create prediction intervals for individual values of the response variable.  

# Required Packages {-}
In the code chunk below, we use the `pacman` package to load the required packages and install them if necessary.

```{r packages}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyverse, 
               magrittr, # for the two-way pipe
               fpp2, # tslm() from forecast,
               astsa, # package is for the required data
               plotly, # for interactive graphs
               stargazer) 
```

# Recap of What we Covered Last Class

## Loading the J&J Data
```{r jj}
jj = jj # assigning the jj data (from astsa) to an object of the same name
class(jj) # to check and see if the class is a ts object
frequency(jj) # to check the frequency --> based on the print out (=4)
jj
p = autoplot(jj) + theme_bw()
ggplotly(p)
```

Based on the plot, we can make three observations:  

  (1) The data is not linear, which means that fitting a linear regression directly to this data is not prudent.  
  (2) The variance was not constant as it increased over time (with larger values of the EPS).  
  (3) The data is exhibiting a seasonal pattern (fourth quarter is consistently below the values for q3 and q1).
  

## Transforming the J&J TS

```{r logtransform}
log_jj = log(jj)
p2 = autoplot(log_jj)
ggplotly(p2) # a linear regression line is probably okay (the variance is more stable with the transformation)
```


## Time as the Independent Variable
```{r extractingTime}
t = time(log_jj) # time makes a decimal date from the ts (if freq  >  1)
t
```

## Fit the Model

```{r model}
model = lm(log_jj ~ t) # t is the ind variable and log_jj is the response
names(model)
summary(model)
```

**Let us recap some of the outputs.** 

### Regression Equation
```{r extractingValuesForEquation}
intercept = summary(model) %>% .[['coefficients']] %>% .['(Intercept)', 'Estimate']
beta1 = summary(model) %>% .[['coefficients']] %>% .['t', 'Estimate']
sigma = summary(model) %>% .[['sigma']]
```
$y_t = `r round(intercept, 4)` + `r round(beta1, 4)`t + \varepsilon, \, \text{where } \varepsilon\sim\mathcal N(0, ~ `r round(sigma, 4)`^2)$ 

### Predicted/Fitted Values and Residuals
```{r fittedVals}
fit = model$fitted.values
res = model$residuals 
```

### Multiple R2
```{r multipleR2, results='asis'}
modelSummary = summary(model)
stargazer(model, type = 'html' )
```



### Tests for Significance
Just talked about the interpretation of the t-statistic for the slope, where we
mentioned that 58.45 means that we are 58.45 standard errors above 0. Hence,
t is a significant predictor of the log_EPS and the p-value is very small.


### Question: What would be our forecast values log EPS for 1981?

#### Approach 1: Manual Calculations

For Q1, my predicted log EPS = `r intercept + beta1*1981.00`.

For Q2, my predicted log EPS = `r intercept + beta1*1981.25`.

For Q3, my predicted log EPS = `r intercept + beta1*1981.50`.

For Q4, my predicted log EPS = `r intercept + beta1*1981.75`.


#### Approach 2: Using the predict or forecast functions

```{r predict1981, results='hold'}
predict_1981 = forecast(
  model, newdata = data.frame(t = c(1981.00, 1981.25, 1981.5, 1981.75) ) )

predict_1981$mean

autoplot(predict_1981)

```




# Use a simple linear regression model for trend adjustment.

Today, we will talk about using the `tslm()`, which comes from the [forecast](https://cran.r-project.org/web/packages/forecast/index.html) in R. 



```{r tslmDiscussion}
model_tslm = tslm(log_jj ~ trend)

model_tslm$data # I am printing out the data used to fit the model
# what we learned from this is
# trend values are integers from 1 to 84 (num obs in time-series)

model_tslm

summary(model_tslm)

# the second nice feature (first being time is coded automatically for us)
# I can use the forecast function directly to predictive h quarters ahead
# this is a function that we have used with other models and it works well
pred_1981_tslm = forecast(model_tslm, h = 4, level = 95)
pred_1981_tslm$mean


pred_1981_1982_tslm = forecast(model_tslm, h = 8, level = 95)
autoplot(pred_1981_1982_tslm) + 
  autolayer( fitted(pred_1981_1982_tslm) ) +
  theme(legend.position = 'bottom')

```

- The intercept of - 0.6678 indicates our estimated value for the log EPS of the J&J stock at time 0 (i.e., Q4 of 1959), i.e., this is meaningful when compared to the intercept from the lm model   

  + because in the lm model time was coded differently 
  
- Let us talk about comparing the slope coefficient in this model (i.e., for trend which is 0.0417) compared to the slope coefficient for the lm model (i.e., for time which was 0.1668)



# Interpret regression diagnostic plots.  

Assumption behind using the forecast function is that we have a model that fits well to our data.


The diagnostics on the regression model confirmed that we do not have an excellent model since our residuals are not iid. Specifically, the:  

- The residuals vary in magnitude as a function of the fitted values.  
- The residuals show a seasonal pattern, where Q2 residuals are typically large and Q4 residuals are typically small.  
- The ACF and PACF show a seasonal pattern in the residuals -- likely need to fit an AR model. 


```{r diagnostics}
source('resPlotTS.R')

residuals_tslm = model_tslm$residuals
fitted_tslm = model_tslm$fitted.values

resplot(res = residuals_tslm, fit = fitted_tslm, freq = 4)

```

# Create prediction intervals for individual values of the response variable.

This is easily obtained based on the output from the `forecast()`. If you wanted to have different intervals then the default of 80 and 95, please change the arguments to level.
