---
title: "28 - Time Series Regression and autoML"
subtitle: "Regression with ARIMA errors and autoML"
author: "Fadel M Megahed"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    paged_df: true
    theme: simplex
    code_folding: show
    code_download: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      verbose = FALSE)
```


# Required Packages
In the code chunk below, we use the `pacman` package to load the required packages and install them if necessary.

```{r packages}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyverse, 
               magrittr, # for the two-way pipe
               fpp2, # accuracy() function, Arima(), auto.arima(),
               astsa, # package is for the required data
               plotly, # for interactive graphs
               stargazer) 

source('resPlotTS.R') # loading our resPlot function from Last Class
```



# Combining Regression with ARIMA Models

We have learned to fit ARIMA models to predict a series from itself, removing the autocorrelation from a series. These methods are useful, but don’t allow us to combine “outside” information to boost our forecast.

Now we will discuss combining regression (**outside information**) and ARIMA models (inside information) to forecast a time series.

A multiple regression model takes on the form:  
$$Y_t = \beta_0 + \beta_1 x_{t,1} + \beta_2 x_{t,2} + \dots + \beta_q x_{t,q} + \epsilon_t,$$
where we typically assume that $\epsilon_t$ is independent, identically distributed white noise (normally distributed).


## What we Learned from the J&J Example 

**Based on the J&J Example**, when our dependent and independent variables are observed over time, $\epsilon_t$  is often correlated over time. **Hence, the assumptions of iid residuals are not met and the fitted regression models should not be used.**

## Possible Solution 

Thus, we will restate our model as follows:  
$$Y_t = \beta_0 + \beta_1 x_{t,1} + \beta_2 x_{t,2} + \dots + \beta_q x_{t,q} + \eta_t,$$ 
where $\eta_t$ **follows an ARIMA model.**  

When we model $\eta_t$, there will be errors from this model, denoted as $\epsilon_t$.  Thus, we have the errors from the regression, $\eta_t$, and the errors from the ARIMA model, denoted as $\epsilon_t$.  Only the errors from the ARIMA, $\eta_t$ are iid white noise.


## Using R to Fit The Model

We will use the “uschange” data from the `fpp2` package to forecast changes in personal consumption expenditures based on personal disposable income from 1970 to 2016.  

**Process:**  

  (1) Start by plotting the quarterly changes in US consumption and personal income  
  (2) Fit a regression with Y = change in consumption and X = Change in personal income, with autocorrelated errors  
  (3) Check the residual plots to ensure the assumptions of the model are met  
  

```{r inclassExample}
# for more information about the data ?fpp2::uschange 
uschange = fpp2::uschange 
uschange

# First step
autoplot(uschange[, 1:2], facet=TRUE)

# Why is the regression model alone inadequate?
reg_regression = lm( uschange[,'Consumption'] ~  uschange[,'Income'] )
summary(reg_regression)

# from the resplot, we conclude that our residuals are NOT iid
# there is significant autocorrelation in the residuals
# we will need an ARIMA type model to account for the correlation in the residuals
resplot(res = reg_regression$residuals,
        fit = reg_regression$fitted.values,
        freq = 4)

# Step2: Fitting a regression model with an ARIMA model on the errors
reg_w_arima_errors = auto.arima(y = uschange[,'Consumption'],
                                xreg = uschange[,'Income'] )

summary(reg_w_arima_errors)
```

From the summary output, our model is:

$y_t = 0.5990 + 0.2028x_t + \eta_t$, where:  

  - $y_t$ is the change in the consumption variable per quarter  
  - $x_t$ is our independent variable (change in income per quarter)
  - $\eta_t = 0.6922\eta_{t-1} + 	\epsilon_t -0.5758\epsilon_{t-1} + 0.1984	\epsilon_{t-2} $


```{r inclassExample2}
resplot(res = reg_w_arima_errors$residuals,
        fit = reg_w_arima_errors$fitted,
        freq = 4)

checkresiduals(reg_w_arima_errors)

```

---

# Things to Explore on your Own

  (1) Compare the results from the class example, with the following three models:  
    (a) auto.arima using only the Y series.  
    (b) autoarima using two explanatory variables ("Income" and "Savings").  
    (c) lm using both income and savings.  
    
  (2) Which models are suitable? (i.e., the assumptions about the model residuals are met).  
  
  (3) Among the suitable models, pick the best model.


```{r activity2}


```


---

# A Very Short Introduction to autoML

In the code chunk below, we will use the h2o package to train and fit multiple machine learning models for predicting the daily closing price of the ziliqa cryptocurrency. For more details on h2o, please refer to the [h2o R Booklet](https://h2o-release.s3.amazonaws.com/h2o/rel-zumbo/1/docs-website/h2o-docs/booklets/RBooklet.pdf)

```{r autoML}
# * Installing the latest release for h2o ----
if(require(h2o) == FALSE) install.packages("h2o")


# * Loading the Other Packages --------------------------------------------
pacman::p_load(tidyquant, # for tq_get and it also loads zoo (package for ts data)
               tidyverse, # for ggplot2, dplyr, etc
               magrittr, # for the two way pipe
               fpp2, # loads the forecast package
               lubridate, # to manipulate "dates"
               DataExplorer, # for plotting missing
               h2o)

zil = tq_get('ZIL-USD', from = '2018-01-01', to = Sys.Date() -1 ) %>% 
  select(date, close, volume)

# * Exploratory Analysis of ZIL Data ----
# (A) Checking to see missing data
colSums(is.na(zil)) # returns the number of missing values per Column
plot_missing(zil, ggtheme = theme_bw(), theme_config = list(legend.position = c('none')))


# * Feature Engineering ----
# (B) Generate some features
df = zil %>%  
  mutate(year = lubridate::year(date),
         quarter = lubridate::quarter(date) %>% as.factor(),
         month = lubridate::month(date, label = TRUE) %>% as.character() %>% as.factor(), # to not make it an ordered factor
         day = lubridate::wday(date, label = T) %>% as.character() %>% as.factor(), # to not make it an ordered factor
         lag1Close = lag(close),
         lag7Close = lag(close, n =7),
         laggedMA7 = rollmeanr(close, k = 7, na.pad = T) %>% lag(),
         laggedMA30 = rollmeanr(close, k = 30, na.pad = T) %>% lag(),
         laggedMA50 = rollmeanr(close, k = 50, na.pad = T) %>% lag(),
         )
str(df) 

df = na.omit(df)

# * Model Fitting ---------------------------------------------------------
# * * Training Data ----
trainData = df %>% filter(year < 2021)
validData = df %>% filter(year == 2021)
testData = df %>% filter(year > 2021)

# * * Response and Predictor Variables
y = "close"
x = setdiff(colnames(df), c("close", "date") )


# * * Using h2o package ----
h2o.init(nthreads = 10)
trainDataH2o = trainData %>% as.h2o()
validDataH2o = validData %>% as.h2o()
testDataH2o = testData %>% as.h2o()

fittedModels = h2o.automl(x = x, 
                          y = y,
                          training_frame = trainDataH2o,
                          validation_frame = validDataH2o,
                          leaderboard_frame = testDataH2o,
                          nfolds = 5,
                          stopping_metric = "AUTO",
                          max_runtime_secs = 120)

# * * Printing the fitted Models
fittedModels

bestModel = fittedModels@leader
bestModel

predictions = h2o.predict(bestModel, newdata = testDataH2o)

testError = h2o.performance(bestModel, newdata = testDataH2o)
testError
```

---


# Exam 04 Information

- Q1 and Q2 interpretation of regression coefficients  

- Q3 interpretation of res plot

- Q4 and Q5 interpretation of tslm() outputs  

- Q6 Interpretation of model summary

- Q7-Q8 interpretation of ARIMA with xreg 

- Q9-Q22 interpretations of which models to fit, autocorrelation, etc based on a plot of a time-series and its ACF 

- Q23-Q32 conceptual multiple choice questions