---
title: "ISA 444: Business Forecasting"
subtitle: "24 - Time Series Regression"
author: Fadel M. Megahed
date: 'Spring 2022'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.8,
                      fig.height= 2.9,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, tidyquant, xtable, magrittr, fpp2, ggpubr)
```

# Preface

### Learning Outcomes for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Explain the simple and multiple linear regression models and interpret the parameters.}
      \item \textbf{Interpret the sample linear regression coefficients in the language of the problem.}
      \item \textbf{Use a simple linear regression model for trend adjustment (time-series data).}
	\end{itemize}
\end{block}



# Simple and Multiple Linear Regression Models

### The Simple Linear Regression Model

$$y_t = \beta_0 + \beta_1 x_{t,1} + \epsilon_t$$ 

$y_t:$

\vspace{0.2in}

$\beta_0:$

\vspace{0.2in}

$\beta_1:$

\vspace{0.2in}

$x_{t,1}:$

\vspace{0.2in}

$\epsilon_t:$

\vspace{0.2in}


### The Multiple Linear Regression Model with $q$ Predictors

$$y_t = \beta_0 + \beta_1 x_{t,1} + \beta_2 x_{t,2} + \dots + \beta_q x_{t,q} + \epsilon_t$$ 


$\beta_0:$

\vspace{0.5in}

$\beta_1, \beta_2, \dots, \beta_q:$

\vspace{0.5in}



# Use a simple linear regression model for trend adjustment (time-series data)

### Live Coding in Class
```{r jj}
pacman::p_load(tidyverse, fpp2, astsa)
jj = jj #from the astsa package
```


 


---

\maketitle