---
title: "ISA 444: Business Forecasting"
subtitle: "06 - Measures of Forecast Accuracy"
author: Fadel M. Megahed
date: 'Spring 2022'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    latex_engine: xelatex
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.95,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = FALSE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, xtable, tidyverse, magrittr)
```

# Preface

### What we Covered Last Class

\begin{block}{\textbf{Main Learning Outcomes}}
$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Understand that \underline{both} charting and numerical summaries of (time-series) data come hand-in-hand} \\
$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Apply transformations to a time series.} 
\end{block}

### Recap: Numerical Summaries do NOT Replace Visuals

\begin{columns}
\begin{column}{0.4\textwidth}
```{r anscombe1, echo = FALSE, results='asis'}
pacman::p_load(tidyverse, tidyquant, Tmisc, xtable) # same data but in 3 columns
df = quartet %>% group_by(set) %>% 
  summarise(x.mean = mean(x), x.sd = sd(x),
            y.mean = mean(y), y.sd = sd(y),
            corr = cor(x, y))
print(xtable(df, align = c(rep('c', 7))), comment = FALSE, size = '\\scriptsize', include.rownames=FALSE)
```
\end{column}%
\begin{column}{0.6\textwidth}
```{r anscombe2, results='asis', echo=FALSE, out.width = '100%'}
ggplot(quartet, aes(x, y)) + geom_point() + 
geom_smooth(method = lm, se = FALSE) + facet_wrap(~set) + theme_bw()
```
\end{column}
\end{columns}



### Recap: Guidelines for Transforming Time-Series Data

\vspace{-\baselineskip}

\begin{figure}
\centering
\adjustbox{max width=\textwidth, frame}{%
\begin{forest}
[\Large{\textbf{Transformation Methods for Time Series}}
[\textbf{Stabilize the Mean}
[\textcolor{miamired}{\textbf{Differencing}}
[Eliminate/Reduce
[Seasonality]
[Trend]]
[Compute New Cases\footnotemark]]]
[\textbf{Stabilize the Variance}
[\textcolor{miamired}{\textbf{Power Transformations}}
[Log]
[Square Root]]] % stabilize the variance
[\textbf{$\sim$Stabilize both}
[\textcolor{miamired}{\textbf{Growth Rates}}]
[\textcolor{miamired}{\textbf{First diff of Log\footnotemark}}
[Exponential Growth TS
[$\ln$ to get a TS with a linear trend]
[Then diff to get a stationary series]
]]]
[\textbf{Rescale}
[\textcolor{miamired}{\textbf{($0-1$) Scaling\footnotemark}}]
[\textcolor{miamired}{\textbf{$z-$transform\footnotemark}}]
]
]
\end{forest}}
\caption{A classification of common transformation approaches for time series data.\footnotemark}
\end{figure}

\vspace{-\baselineskip}

\addtocounter{footnote}{-5}

\stepcounter{footnote}\footnotetext{The \href{https://covid19datahub.io/}{COVID19 package} returns cumulative cases, i.e. a first difference $\longrightarrow$ new confirmed cases.}

\stepcounter{footnote}\footnotetext{First difference of LOG $\approxeq$ percentage change. This is almost exact if the percentage change is small, but for larger percentage changes, it may differ greatly (see \href{https://faculty.fuqua.duke.edu/~rnau/Decision411_2007/411log.htm}{here for more details}).}

\stepcounter{footnote}\footnotetext{Rescaling of the data from the original range so that all values are within the range of 0 and 1. Mathematically, speaking this can be achieved by calculating $y_t = \frac{x_t - \min}{\max - \min}$.}

\stepcounter{footnote}\footnotetext{One can normalize a time-series by $z_t = \frac{x_t - \mu}{\sigma}$.}

\stepcounter{footnote}\footnotetext{My (incomplete) attempt to provide you with a taxonomy for time series data transformations.}





### Learning Objectives for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
\begin{itemize}
\item \textbf{Apply and interpret measures of forecast accuracy.} 
\item \textbf{Interpret prediction intervals for a simple forecast.}
\end{itemize}
\end{block}



# Measures of Forecast Accuracy

### Recap: Definition of Forecast

\begin{figure}
\centering
\href{https://tinyurl.com/y5h5k4aj}{\includegraphics[width=0.9\textwidth, height = 0.7\textheight, frame, keepaspectratio]{Figures/forecastDef.png}}
\caption{The definition of the term "forecast" as obtained from Bing/Merriam-Webster.}
\end{figure}


### A Naive Forecast

- A naïve forecast for an observation, $Y_t$ , is the observation prior, $Y_{t-1}$.  

- For some types of time series (e.g. Random Walks), a naïve forecast is
the best possible forecast one can make\footnotemark.

- In the case of seasonal data, a naïve forecast could be the observation
from the prior period.   
- For example, in the case of monthly data, the naïve forecast for the observation $Y_{Jan 2018}$ could be $Y_{Jan 2017}$. In this case, we would denote the frequency, m=12, and the naïve forecast for $Y_t$ is the observation $m$ periods prior, or $Y_{t-m}$.

\footnotetext{Slide is from \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}


### Measures of Forecast Accuracy

The measures of accuracy we will discuss all deal with the difference
between the actual observed value ($Y_t$) and the forecasted value ($F_t$) at
time $t$. In order to measure forecast accuracy, we assume we have $m$ actual
values available, thus we have $Y_{t+1}, \, Y_{t+2}, \, \dots, \, Y_{t+m}$ and forecasts $F_{t+1}, \, F_{t+2}, \, \dots, \, F_{t+m}$. This is important because we will be averaging the forecast errors over $m$.\footnotemark

\footnotetext{Note that your reference book discusses rolling forecast origins. This is important, but we will save this discussion for later in the semester. For now, assume the forecast origin is fixed (i.e., we are only interested in the one-period ahead forecast).}


**Forecast Error:** 
\begin{equation}
e_{t+i} = Y_{t+i} - F_{t+i}.
\end{equation}

### Measures Reflecting "Average" Forecast Performance

**Mean Error:** 
\begin{equation}
ME = \frac{\sum_{i=1}^{m} e_{t+i}}{m}.
\end{equation}

**Mean Percentage Error:** 
\begin{equation}
MPE = \frac{100}{m}\sum_{i=1}^{m}\frac{ e_{t+i}}{Y_{t+i}}.
\end{equation}


### Computing Measures of "Average" Forecast Performance

```{r meanError, echo=FALSE, results="asis"}
pacman::p_load(tidyquant, magrittr)
cardano = tq_get("ADA-USD", from = "2022-02-01", to = "2022-02-07") %>% 
  select(symbol, date, adjusted)
cardano %<>% mutate(naiveFC = lag(adjusted),
                    e = adjusted - naiveFC,
                    PE = 100*e/adjusted)
cardano$date %<>%  as.character()
print(xtable(cardano , align = c(rep('c', 7))), comment = FALSE, size = '\\scriptsize', include.rownames=FALSE)
cat(paste0("The ME and MPE are equal to ", mean(cardano$e, na.rm = T) %>% round(2), " and ", mean(cardano$PE, na.rm = T) %>% round(2),
           "% , respectively."))
```

**Comments:**     

- We are forecasting the adjusted/closing price of the $ADA stock. Given that this crypto, trading is 24/7 (i.e., the **next trading day always equals the next day**).    
- The naïve forcast is the `lag(1)` of the series; thus, the forecast error is the `diff(1)`.  
- A **seasonal naive forecast** can be computed using `lag(m)` or the `snaive()` from the [fpp2 package](https://cran.r-project.org/web/packages/fpp2/fpp2.pdf).  


### Measures Reflecting "Variablity" in Forecast Performance

**Absolute Forecast Error:** 
\begin{equation}
|e_{t+i}| = |Y_{t+i} - F_{t+i}|.
\end{equation}

**Squared Forecast Error:** 
\begin{equation}
(e_{t+i})^2 = (Y_{t+i} - F_{t+i})^2.
\end{equation}

**Mean Absolute Error:** 
\begin{equation}
MAE = \frac{\sum_{i=1}^{m}|e_{t+i}|}{m}.
\end{equation}


**Root Mean Squared Error:** 
\begin{equation}
RMSE = \sqrt{\frac{\sum_{i=1}^{m}(e_{t+i})^2}{m}}.
\end{equation}


### Measures Reflecting "Relative" Forecast Performance

**Mean Absolute Percentage Error:** 
\begin{equation}
MAPE = \frac{100}{m} \sum_{i=1}^{m} \frac{|e_{t+i}|}{m}.
\end{equation}

**Relative Mean Absolute Error:**
\begin{equation}
RelMAE = \frac{\sum_{i=1}^{m}|e_{t+i}|}{\sum_{i=1}^{m} |Y_{t+i} - Y_{t+i-1}|}.
\end{equation}

**Thiel's U:**
\begin{equation}
U = \sqrt{ \frac{\sum_{i=1}^{m}(e_{t+i})^2}{\sum_{i=1}^{m} (Y_{t+i} - Y_{t+i-1})^2} }.
\end{equation}


### An Overview of Computing these Measures in R {.allowframebreaks}

```{r computations1}
pacman::p_load(tidyquant, magrittr, fpp2, xtable)

cardano = tq_get("ADA-USD", from = "2022-02-01", to = "2022-02-07") %>% 
  select(symbol, date, adjusted)
cardano %<>% mutate(naiveFC = lag(adjusted))

e = cardano$adjusted - cardano$naiveFC
ME = mean(e, na.rm=T)
RMSE = mean(e^2, na.rm=T) %>% sqrt()
MAE = abs(e) %>% mean(na.rm=T)
MPE = 100 * mean(e/cardano$adjusted, na.rm=T)
MAPE= 100 * mean(abs(e)/cardano$adjusted, na.rm=T)
```


```{r computations2}
E = c(ME, RMSE, MAE, MPE, MAPE)
names(E) = c("ME", "RMSE", "MAE", "MPE", "MAPE")
round(E, 2) %>% print()

# Alternatively, we could have just computed it using the fpp2 package
accuracy(object = cardano$naiveFC, # forecast object is the first argument
         cardano$adjusted) %>% round(2)
```


# Prediction Intervals

### Point vs Interval Forecasts

- **Point Forecasts:** future observations for which we report a single forecast observation.  

- **Interval Forecast:** a range of values that are reported to forecast an outcome.


If we assume the forecast errors follow a Normal Distribution, an approximate $100(1-\alpha)$ prediction interval can be computed as follows:  
\begin{equation}
\hat{F}_t \pm Z^**RMSE,
\end{equation}
where:  

- $\hat{F}_t$ forecast at time $t$.  
- The RMSE can be used as an estimate of the standard deviation of the forecast errors.  
- $Z^*$ is the quantile corresponding to $100(1-\frac{\alpha}{2})$.


### Recall: Standard Normal Distribution [1]

```{r normPlot, echo=FALSE, fig.height= 2.9}
prob = 0.95

mean = 0
sd = 1

ll = qnorm((1-prob)/2)
ul = qnorm(prob + ((1-prob)/2))

normal_data = data.frame(
  `z_score`= mean + seq(-4,4,length=100)*sd,
  Density = dnorm(x = seq(-4,4,length=100), mean, sd)
)

# Polygon and its calculations are based on 
# https://rstudio-pubs-static.s3.amazonaws.com/198507_702e146703914ebab1e5bf8e48a323aa.html

color_data_seq = seq(from = ll, to = ul, by = .01)
color_data = data.frame(x = color_data_seq, y = dnorm(color_data_seq))
color_data = rbind(c(min(color_data$x),0),
                   color_data,
                   c(max(color_data$x),0))

left_tail_seq = seq(from = -4, to = ll, by = .01)
left_tail = data.frame(x = left_tail_seq, y = dnorm(left_tail_seq))
left_tail = rbind(c(min(left_tail$x),0),
                  left_tail,
                  c(max(left_tail$x),0))

normal_data %>% 
  ggplot(aes(x = `z_score`, y = Density)) +
  geom_line() +
  # middle area
  geom_segment(aes(x = ll, xend = ll, y=0, yend = dnorm(ll,0,1)), color = 'red', size = 2) +
  geom_segment(aes(x = ul, xend = ul, y=0, yend = dnorm(ul,0,1)), color = 'red', size = 2) +
  geom_polygon(data = color_data, aes(x, y), fill = 'red', alpha = 0.2) +
  geom_text(x = 0, y = 0.2, size = 3, fontface = "bold",
            label = paste0(prob, '\n between both \n red lines'), color = 'red') +
  geom_text(x = ul+ 0.5, y = dnorm(ul)+0.05, size = 3, fontface = "bold", 
            label = paste0('Z* = ', round(qnorm(prob + (1-prob)/2), 2)), 
            color = 'red') +
  geom_text(x = 0.25, y = 0.05+(dnorm(ul)/2), size = 3, fontface = 'bold',
            label = paste0('Z* computed as', 
                           '\n = qnorm(', prob, '+', (1-prob)/2, ')'),
            color = 'black') +
  # arrows
  geom_segment(aes(x = ul + 0.25, xend = ul, y=dnorm(ul)+0.03, 
                   yend = dnorm(ul,0,1)), color = 'red', 
               size = 1, arrow = arrow(length = unit(0.3,"cm"))) +
  geom_segment(aes(x = 0.25, xend = ul, y=dnorm(ul)/2, 
                   yend = dnorm(ul)/2), color = 'black', 
               size = 1, arrow = arrow(length = unit(0.3,"cm"))) +
  # left_tail_area
  geom_polygon(data = left_tail, aes(x, y), fill = 'yellow', alpha = 0.3) +
  geom_text(x = (-4 - ll/2), y = 0.025, size = 3, fontface = "bold",
            label = paste0(((1-prob)/2)), color = 'black') +
  scale_y_continuous(limits = c(0,0.42)) +
  theme_bw() +
  labs(caption = "See http://rstudio.fsb.miamioh.edu:3838/megahefm/isa444/class06/ for an interactive version of this plot")
```



### Prediction Intervals for the $ADA Data {.allowframebreaks}

```{r pIs}
naiveFC = cardano$naiveFC
PInormU = naiveFC + abs(qnorm(0.975))*RMSE
PInormL = naiveFC - abs(qnorm(0.975))*RMSE
dfNaive = data.frame(date = cardano$date, Ft = naiveFC, PInormU, PInormL)
dfNaive %>% ggplot(aes(x= date, y = naiveFC)) + geom_line() +
  theme_bw() + labs(title ="Overlaying 95% PIs in ggplot", fill = "95% PI") +
  geom_ribbon(aes(ymin = PInormL , ymax = PInormU, fill = "band"), 
              alpha = 0.2, color = "red") + theme(legend.position= "bottom")
```

# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
\begin{itemize}
\item \textbf{Apply transformations to a time series.}
\item \textbf{Apply and interpret measures of forecast accuracy.}
\item \textbf{Interpret prediction intervals for a simple forecast.}
\end{itemize}
\end{block}


### Things to Do

- **Recommended:** Thoroughly read [Chapter 2 of our reference book](https://wessexlearning.com/products/principles-of-business-forecasting-2nd-ed-part-i).  

- Go through the slides, examples and make sure you have a good understanding of what we have covered.  
- **Potential Practice Problems:**   

- Extract data using either the `tq_get()` ([tidyquant package](https://business-science.github.io/tidyquant/)) or the `covid19()` ([COVID19 package](https://covid19datahub.io/)), and compute the transformations using a manual (i.e., Excel) approach and R. Then, interpret the obtained transformed series.   
- **Reference Book Example:** For the Means approaches in Example 2.7 (P.49), use R to compute the 7 error forecasting metrics (data available [here](https://www.wessexlearning.org/pobf2e/index.html)).    
- **Reference Book Exercise 2.12:** Compute the forecast errors for the naive forecast.

- **Required:** Complete the [graded assignment](https://miamioh.instructure.com/courses/167327/quizzes/449739).


### Graded Assignment 05: Evaluating your Understanding

Please go to \href{https://miamioh.instructure.com/courses/167327/quizzes/449739}{Canvas (click here)} and answer the five questions. **Due September 14, 2021 [8:30 AM].** 

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of the material covered up to the end of Class 06. To reinforce your understanding of the covered material, I also suggest reading [all of Chapter 2 of the reference book](https://wessexlearning.com/products/principles-of-business-forecasting-2nd-ed-part-i).

**General Guidelines:**  

\vspace{-0.5\baselineskip}

- Individual assignment.  
- This is **NOT** a timed assignment.    
- Proctorio is NOT required for this assignment. 


---

\maketitle