---
title: "ISA 444: Business Forecasting"
subtitle: "16 - Autocorrelation and Partial Autocorrelation"
author: Fadel M. Megahed
date: 'Spring 2022'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    latex_engine: xelatex
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.95,
                      fig.height= 2.5,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = FALSE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(table, tidyverse, magrittr, tidyquant, fpp2, scales, sweep, timetk, readxl)
library(ggplot2); theme_set(theme_bw(base_size = 8))
```

# Preface

### What we Covered Last Class
\begin{block}{\textbf{Main Learning Outcomes}}
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define the population mean, and variance of a random variable.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define the population covariance, and correlation between two random variables.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Define the population autocovariance and autocorrelation of a random variable.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Use sample estimates of the population mean, variance, covariance, and correlation.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Explain the properties of the large sample distribution of the sample ACF.} \\
			$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{Use the large sample distribution of the sample ACF to identify significant autocorrelation in a time series.} \\
\end{block}


### Recap: ACF Definitions
We defined **autocorrelation as the correlation between a time-series and its past values.**  

The **ACF was defined as a function that captures the correlations between pairs of values at a certain lag**. For example:  

- Lag-1 autocorrelation captures the correlation between $y_t$ and $y_{t-1}$.  
- Lag-2 autocorrelation captures the correlation between $y_t$ and $y_{t-2}$. 

### Recap: HW 09 - Autocorrelation

Per your feedback post exam 01, let us discuss the solutions to [HW 09](https://miamioh.instructure.com/courses/167327/quizzes/449736) over the next few minutes.

**As usual**, please do not hesitate to ask me questions.


### Recap: The ACF of Advance Retail Sales [RSCCAS]

Build on the example below and plot the ACF for:  

- The [RSCCAS](https://fred.stlouisfed.org/series/RSCCAS) series, which is stored in the `retail` df.  

- The first differences of the [RSCCAS](https://fred.stlouisfed.org/series/RSCCAS) series.  

```{r retailSales}
pacman::p_load(tidyverse, magrittr, fpp2, tidyquant, scales, ggpubr)
retail = tq_get('RSCCAS', from = ymd('1992-01-01'), to = Sys.Date(),
                get = 'economic.data')
```

\textcolor{miamired}{\textbf{Based on both plots, answer the following two questions:}}
\begin{itemize}
  \item \textcolor{miamired}{Are the limits for both ACF plots the same? Why? Why not?}
  \item \textcolor{miamired}{Comment on how their ACF plots are different.}
\end{itemize}

```{r retailSales2, echo=F, eval = F}
retail %<>% mutate(diff = price - lag(price), 
                   logSales = log(price),
                   diffLogSales = logSales - lag(logSales))
p1 = retail %>% ggplot(aes(x = DATE, y= price)) + geom_line()
p2 = acf(retail$price, plot = F) %>% autoplot()
p3 = acf(retail$diff, plot = F, na.action = na.pass) %>% autoplot()
p4 = acf(retail$diffLogSales, plot = F, na.action = na.pass) %>% autoplot()

ggarrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

### Learning Objectives for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Determine if a sample ACF plot “cuts off” or “dies down”.}
			\item \textbf{Explain how sample partial autocorrelation is calculated.}
	\end{itemize}
\end{block}


# Explaining Sample ACF Plots

### Some Time Series and their ACF Plots [1]

```{r simMA1, echo=FALSE}
#Generate MA processes of order 2
set.seed(20210309)
ma.sim1 = arima.sim(model=list(ma=c(.4,.9)),n=100)
p1 = ma.sim1 %>% autoplot() + labs(y = 'Y')
p2 = acf(ma.sim1, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2,
          labels = c('MA(2)', ''),
           ncol = 1, nrow = 2)
```


### Some Time Series and their ACF Plots [2]
```{r simMA2, echo=FALSE}
set.seed(20210310)
ma.sim2 = arima.sim(model=list(ma=c(.4,.6,.7)),n=100)
p1 = ma.sim2 %>% autoplot() + labs(y = 'Y')
p2 = acf(ma.sim2, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2,
          labels = c('MA(3)', ''),
           ncol = 1, nrow = 2)
```


### Some Time Series and their ACF Plots [3]

```{r simAR1, echo=FALSE}
set.seed(20210310)
ar.sim1 = arima.sim(model=list(ar=c(.9)),n=500)
p1 = ar.sim1 %>% autoplot() + labs(y = 'Y')
p2 = acf(ar.sim1, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2,
          labels = c('AR(1): 0.9', ''),
           ncol = 1, nrow = 2)
```


### Some Time Series and their ACF Plots [4]

```{r simAR1-2, echo=FALSE}
set.seed(20210310)
ar.sim2 = arima.sim(model=list(ar=c(-.9)),n=500)
p1 = ar.sim2 %>% autoplot() + labs(y = 'Y')
p2 = acf(ar.sim2, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2,
          labels = c('AR(1): -0.9', ''),
           ncol = 1, nrow = 2)
```


### Some Time Series and their ACF Plots [5]

```{r google, echo=FALSE}
GOOG = getSymbols('GOOG', from = '2019-01-01', to = '2020-12-31', auto.assign = FALSE)
GOOG = GOOG$GOOG.Adjusted
p1 = GOOG %>% autoplot()
p2 = acf(GOOG, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, nrow = 2)
```

# Partial Autocorrelation


### Definition: General

**Statistical Definition:** Let us say that we have three variables, $X$, $Y$, and $Z$, all correlated, and we want to know how $X$ and $Y$ are correlated after we remove the effects of $Z$ on each.


**Computation Approach:**

\begin{center}
$\hat{X} = a_1 + b_1Z; \qquad \qquad X^* = X - \hat{X}$ 

$\hat{Y} = a_2 + b_2Z; \qquad \qquad Y^* = Y - \hat{Y}$ 
\end{center}


$Corr(X^*, Y^*)$ is the partial correlation between $X$ and $Y$. It is the correlation that remains after we remove the effect of $Z$.


### PACF in the Context of Time-Series Analysis
The Partial Autocorrelation between $Y_t$ and $Y_{t+k}$ is the correlation between $Y_t$ and $Y_{t+k}$ after removing the effects of $Y_{t+1}, \, Y_{t+2}, \, Y_{t+3}, \, \dots, \, Y_{t+k-1}$.  

  - We plot the partial autocorrelation over multiple lags just like the autocorrelation function (ACF).  
  - We refer to the plotted partial autocorrelations as the PACF.


### Some Time Series and their PACF Plots [1]
```{r maSim1PACF, echo=FALSE}
p1 = ma.sim1 %>% autoplot() + labs(y = 'Y')
p2 = acf(ma.sim1, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)
p3 = pacf(ma.sim1, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2, p3,
          labels = c('MA(2)', '', ''),
           ncol = 1, nrow = 3)
```


### Some Time Series and their PACF Plots [2]
```{r maSim2PACF, echo=FALSE}
p1 = ma.sim2 %>% autoplot() + labs(y = 'Y')
p2 = acf(ma.sim2, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)
p3 = pacf(ma.sim2, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2, p3,
          labels = c('MA(3)', '', ''),
           ncol = 1, nrow = 3)
```

### Some Time Series and their PACF Plots [3]
```{r arSim1PACF, echo=FALSE}
p1 = ar.sim1 %>% autoplot() + labs(y = 'Y')
p2 = acf(ar.sim1, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)
p3 = pacf(ar.sim1, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2, p3,
          labels = c('AR(1): 0.9', '', ''),
           ncol = 1, nrow = 3)
```


### Some Time Series and their PACF Plots [4]
```{r arSim2PACF, echo=FALSE}
p1 = ar.sim2 %>% autoplot() + labs(y = 'Y')
p2 = acf(ar.sim2, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)
p3 = pacf(ar.sim2, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2, p3,
          labels = c('AR(1): -0.9', '', ''),
           ncol = 1, nrow = 3)
```


### Some Time Series and their PACF Plots [5]

```{r googlePACF, echo=FALSE}
p1 = GOOG %>% autoplot()
p2 = acf(GOOG, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)
p3 = pacf(GOOG, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p1, p2, p3, ncol = 1, nrow = 3) 
```


### What we have learned from the GOOG Example [1]

```{r GOOGscatter, echo=FALSE}
google = data.frame(adjusted = as.vector(GOOG) %>% round(0), 
                    lag1price = as.vector(GOOG) %>% lag() %>% round(0))
google %>% ggplot(aes(x = lag1price, y = adjusted)) +
  geom_point() + theme_bw() 
```

### What we have learned from the GOOG Example [2]
```{r googleLM, echo=FALSE, results='asis'}
model = lm(data = google, adjusted ~ lag1price) 
stargazer::stargazer(model, font.size = 'scriptsize', header = FALSE)
```


### What we have learned from the GOOG Example [3]
```{r googleDiff1, echo=FALSE}
googleDiff = c(NA, diff(google$adjusted))  
googleDiff = xts(x = googleDiff, order.by = date(GOOG)) %>% na.omit()
googleDiff %>% autoplot() + theme_bw() 
```

### What we have learned from the GOOG Example [4]
```{r googleDiff2, echo=FALSE}
p2 = acf(googleDiff, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)
p3 = pacf(googleDiff, plot = FALSE, lag.max = 12) %>% 
  autoplot() + labs(title = NULL)

ggarrange(p2, p3, ncol = 1, nrow = 2)
```

# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Determine if a sample ACF plot “cuts off” or “dies down”.}
			\item \textbf{Explain how sample partial autocorrelation is calculated.}
	\end{itemize}
\end{block}



### Things to Do
 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered.  
 - Go through the [self-paced tutorial for week 08](http://rstudio.fsb.miamioh.edu:3838/megahefm/isa444/week08/) (which will take you sometime between 5-10 minutes to complete).  
 
 - Complete the assignment (see details in next slide).  
 

### Graded Assignment 10: Evaluating your Understanding

Please go to \href{https://miamioh.instructure.com/courses/167327/quizzes/449730?module_item_id=3678615}{Canvas (click here)} and answer the questions. **The assignment is due March 18, 2022 [8:30 AM].**

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of autocorrelation and PACF. To reinforce your understanding of the covered material, I also suggest going through the [self-paced tutorial for week 08](http://rstudio.fsb.miamioh.edu:3838/megahefm/isa444/week08/) (which will take you sometime between 5-10 minutes to complete).

**General Guidelines:**  

\vspace{-0.5\baselineskip}

  - Individual assignment.  
  - This is **NOT** a timed assignment.    
  - Proctorio is NOT required for this assignment.  
  - You will need to have R installed (or accessible through the \href{https://virtualpc.fsb.miamioh.edu/RDWeb/Pages/en-US/default.aspx}{Remote Desktop})



---

\maketitle