---
title: "ISA 444: Business Forecasting"
subtitle: "18 - ARMA Models"
author: Fadel M. Megahed
date: 'Spring 2022'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    latex_engine: xelatex
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.8,
                      fig.height= 2.9,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, tidyquant, xtable, magrittr, fpp2, ggpubr)
library(ggplot2); theme_set(theme_bw(base_size = 8))
```

# Preface


### Quick Refresher on What we Covered in Class so Far {.allowframebreaks}

\begin{block}{\textbf{Main Learning Outcomes Discussed in Class so Far}}
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have studied the \textcolor{miamired}{basic components of time series}, including trends, seasonal components, and cyclical components.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have considered \textcolor{miamired}{why and how} we forecast.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have learned how we \textcolor{miamired}{evaluate forecast accuracy} with measures such as Mean Absolute Error, Root Mean Square Error, Mean Absolute Percent Error.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have learned to compute forecast accuracy to evaluate both \textcolor{miamired}{in-sample} and \textcolor{miamired}{out-of-sample performance} of forecasts.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have considered \textcolor{miamired}{Moving Average, Decomposition,} and \textcolor{miamired}{Smoothing Methods} for exploring and forecasting time series.} 
\end{block}


\begin{block}{\textbf{Main Learning Outcomes Discussed in Class so Far}}
  $\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have learned \textcolor{miamired}{preliminaries} regarding the \textcolor{miamired}{autocorrelation structure} of time series and how to plot the \textcolor{miamired}{autocorrelation} and \textcolor{miamired}{partial autocorrelation} over time.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We have studied a \textcolor{miamired}{random walk model} and know how to recognize one using the ACF function.} \\
		$\quad$ \textcolor{darkgreen}{\large \checkboxFadel} \textbf{We know what it means for a time series to be \textcolor{miamired}{nonstationary}, and how to test for this formally.} \\
\end{block}


### Where We are Going
  - Using all the information we have learned, we will learn to **Formally Model** a time series with statistical models.  

  - Some of these models will be **Extrapolative**, and some will be **Causal**.


### Learning Objectives for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Describe the behavior of the ACF and PACF of an AR(p) process.}
			\item \textbf{Describe the behavior of the ACF and PACF of an MA(q) process.}
			\item \textbf{Describe the behavior of the ACF and PACF of an ARMA (p,q) process.}
			\item \textbf{Fit an ARMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARIMA model.}
	\end{itemize}
\end{block}


### Preface: ARMA Models

Models we consider here may have two components, an autoregressive component (AR) and a moving average component (MA).


# Autoregressive Processes

### The First Order Autoregressive Process {.allowframebreaks}

The **First Order Autoregressive Process—AR(1)** is given by  
\begin{equation*}
y_t = \delta + \phi y_{t-1} + \epsilon_t,
\end{equation*}
where $|\phi| < 1$ is a weight, and $\epsilon_t$ is white noise. Essentially, this is similar (not exactly the same though) as regressing $y_t$ on $y_{t-1}$. 

\begin{equation*}
E(y_t) = \mu = \frac{\delta}{1 - \phi}
\end{equation*}

\begin{equation*}
Var(y_t) = \sigma^2 \frac{1}{1 - \phi^2}
\end{equation*}

The *population* autocorrelation function of the AR(1) process at lag $k$ is
\begin{equation*}
\rho(k) = \phi^k
\end{equation*}


The theoretical/population ACF of an AR(1) process with $\phi = 0.6$ will look like this:

```{r theoraticalacf, fig.height=1.75, echo=FALSE}
arTheoryACF = data.frame(lag = 0:12)
arTheoryACF %<>% mutate(acf = 0.6^lag)
ggplot(arTheoryACF, aes(x = lag, y = acf)) + geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = scales::pretty_breaks(n=13)) + xlim(0, 13) +
  theme_bw()
```

### Example Plots of Simulated AR(1) Data [1]

```{r ar1sim1, echo=FALSE}
ar.sim = arima.sim(model=list(ar=c(.8)),n=500)

p1 = ar.sim %>% autoplot() + theme_bw() + labs(y = 'AR(1): 0.8')
p2 = acf(ar.sim, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### Example Plots of Simulated AR(1) Data [2]

```{r ar1sim2, echo=FALSE}
ar.sim = arima.sim(model=list(ar=c(-.8)),n=500)

p1 = ar.sim %>% autoplot() + theme_bw() + labs(y = 'AR(1): -0.8')
p2 = acf(ar.sim, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL) + ylim(c(-1,1))

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### Example Plots of Simulated AR(1) Data [3]

**Notice how the ACF “dies down” in each case.**


### The Second Order Autoregressive Process {.allowframebreaks}

The **Second Order Autoregressive Process—AR(2)** is given by  
\begin{equation*}
y_t = \delta + \phi_1 y_{t-1} \color{miamired}{+ \phi_2 y_{t-2} }  \color{black}{+ \epsilon_t},
\end{equation*}
where $|\phi_1| < 1 \text{ and } |\phi_2| < 1$ are weights, and $\epsilon_t$ is white noise. Essentially, this is similar (not exactly the same though) as regressing $y_t$ on $y_{t-1} \text{ and } y_{t-2}$. 


\begin{equation*}
E(y_t) = \mu = \frac{\delta}{1 - \phi_1 - \phi_2}
\end{equation*}

\begin{equation*}
Var(y_t) = \phi_1 \gamma(1) + \phi_2 \gamma(2) + \sigma^2,
\end{equation*}
where $\gamma(1)$ and $\gamma(2)$ are the autocovariance functions at lags 1 and 2, respectively.

The *population* autocorrelation function of the AR(1) process at lag $k$ is
\begin{equation*}
\rho(k) = \phi_1 \rho(k-1) + \phi_2 \rho(k-2)
\end{equation*}

The AR(2) model can be seen as an “adjusted” AR(1) model for which a single exponential decay expression as in the AR(1) model is not enough to describe the pattern in the ACF. Hence an additional term for the
second lag is added.



### Example Plots of Simulated AR(2) Data [1]

```{r ar2sim1, echo=FALSE}
ar.sim1 = arima.sim(model=list(ar=c(0.4, 0.5)), n=500)

p1 = ar.sim1 %>% autoplot() + theme_bw() + labs(y = 'AR(2): 0.4, 0.5')
p2 = acf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### Example Plots of Simulated AR(2) Data [2]

```{r ar2sim2, echo=FALSE}
ar.sim2 = arima.sim(model=list(ar=c(-0.4, 0.5)),n=500)

p1 = ar.sim2 %>% autoplot() + theme_bw() + labs(y = 'AR(2): -0.4, 0.5')
p2 = acf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL) + ylim(c(-1,1))

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### The General Order Autoregressive Process—AR(p) {.allowframebreaks}

The **General Order Autoregressive Process—AR(p)** is given by
\begin{equation*}
y_t = \delta + \phi_1 y_{t-1} \color{miamired}{+ \phi_2 y_{t-2} + \dots + \phi_p y_{t-p}}  \color{black}{+ \epsilon_t},
\end{equation*}
where $|\phi_i| < 1 \, \forall i = 1, 2, \dots , p$ are weights, and $\epsilon_t$ is white noise. Essentially, this is similar (not exactly the same though) as regressing $y_t$ on $y_{t-1}, \dots, y_{t-p}$. 


\begin{equation*}
E(y_t) = \mu = \frac{\delta}{1 - \phi_1 - \phi_2 - \dots - \phi_p}
\end{equation*}

\begin{equation*}
Var(y_t) = \sum_{i=1}^{p}\phi_i \gamma(i) +  \sigma^2,
\end{equation*}
where $\gamma(i)$ is the autocovariance functions at lag $i$.

The *population* autocorrelation function of the AR(2) process at lag $k$ is
\begin{equation*}
\rho(k) = \sum_{i=1}^{p}\phi_i \rho(k-i) \text{ for } k > 0
\end{equation*}


The ACF of an AR($p$) process, for $p>1$ is a mixture of exponential decay and a damped sinusoid expression (damped sinusoid from the lag 2 and greater).

### AR Model: Determining if the Data Can Be Modeled as an AR Process

  - We can usually tell from the ACF that there is an autoregressive (AR) component to the data because the ACF plot tends to geometrically decrease in magnitude (i.e., "die down").  
  
  - The **Order** of an AR Process refers to how many lags you include in the autoregressive model.  
  
  - Because the ACF of the AR model is a mixture, the **ACF is not useful for determining the order of the AR process**.  

  - Thus, the ACF helps us to know that we have an **AR model**, but not which AR model to fit!  
  

### AR Model: Determining the Order

Recall the **Partial Autocorrelation:** The Partial Autocorrelation between $y_t$ and $y_{t+k}$ is the correlation between $y_t$ and $y_{t+k}$ removing the effects of $y_{t+1}, y_{t+2}, \dots, y_{t+k-1}$.  

  - When plotted over multiple lags, we refer to the plot as the Partial Autocorrelation Function or PACF.  
  
  - For an AR($p$) model, the PACF between $y_t$ and $y_{t+k}$ should be 0 $\forall k > p$.  
  
  - Thus, for an AR($p$) process, the PACF should “cut off” after lag $p$.



### Example Plots of AR Processes [1]

```{r ar2sim1b, echo=FALSE}
p1 = ar.sim1 %>% autoplot() + theme_bw() + labs(y = 'AR(2)')
p2 = acf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
p3 = pacf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, p3, ncol = 1, 
          labels = c('AR(2): 0.4, 0.5', '', ''), nrow = 3) 
```

### Example Plots of AR Processes [2]

```{r ar2sim2b, echo=FALSE}
p1 = ar.sim2 %>% autoplot() + theme_bw() +labs(y = 'AR(2)')
p2 = acf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL) + ylim(c(-1,1))
p3 = pacf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, p3, ncol = 1, 
          labels = c('AR(2): -0.4, 0.5', '', ''), nrow = 3) 
```

### Example Plots of AR Processes [3]

```{r ar3sim, echo=FALSE}

ar.sim3 = arima.sim(model=list(ar=c(0.4, -0.5, 0.3)),n=500)

p1 = ar.sim3 %>% autoplot() + theme_bw() +labs(y = 'AR(3)')
p2 = acf(ar.sim3, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
p3 = pacf(ar.sim3, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, p3, ncol = 1, 
          labels = c('AR(3): 0.4, -0.5, 0.3', '', ''), nrow = 3) 
```


### Example Plots of AR Processes [4]
**Note how the PACF is not significant after lag (p) in each case.**


# The Moving Average (MA) Process


### The Moving Average Process {.allowframebreaks}

The moving average process of order $q, MA(q)$, process is given as
\begin{equation*}
  y_t = \mu + \epsilon_t - \theta_1 \epsilon_{t-1}  - \dots - \theta_q\epsilon_{t-q} 
\end{equation*}
where $\theta_i$ is a weight, and $\epsilon_i$ is white noise. **An MA(q) process is always stationary regardless of the weights.**

\begin{equation*}
  \begin{split}
    E(y_t) &= E(\mu + \epsilon_t - \theta_1 \epsilon_{t-1}  - \dots - \theta_q\epsilon_{t-q}) \\
           &= \mu
  \end{split}
\end{equation*}

\begin{equation*}
  \begin{split}
    Var(y_t) &= Var(\mu + \epsilon_t - \theta_1 \epsilon_{t-1}  - \dots - \theta_q\epsilon_{t-q}) \\
           &= \sigma^2 (1 + \theta_1^2 + \dots + \theta_q^2)
  \end{split}
\end{equation*}

The POPULATION autocorrelation function of the MA($q$) process at lag $k$ is

\begin{equation*}
  \rho(k) = 
  \begin{cases}
    \frac{(-\theta_k + \theta_1 \theta_{k+1} + \dots + \theta_{q-k} \theta_{q})}{1+ \theta_1^2 + \dots + \theta_q^2}, & k = 1, \, 2, \dots, \, q \\
    0, & k > q
  \end{cases}
\end{equation*}

This feature of the ACF is very helpful in identifying the MA model and its appropriate order because the ACF function of a MA model is not significant (i.e., “cuts off”) after lag $q$.



### Example Plots of MA Processes [1]

```{r ma1sim, echo=FALSE}
ma.sim = arima.sim(model=list(ma=c(0.7, 0.5)),n=500)

p1 = ma.sim  %>% autoplot() + theme_bw() +labs(y = 'MA(2): 0.7, 0.5')
p2 = acf(ma.sim, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### Example Plots of MA Processes [2]

```{r ma2sim, echo=FALSE}
ma.sim = arima.sim(model=list(ma=c(0.7, 0.5, 0.7)),n=500)

p1 = ma.sim  %>% autoplot() + theme_bw() +labs(y = 'MA(2): 0.7, 0.5, 0.7')
p2 = acf(ma.sim, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### Example Plots of MA Processes [3]

```{r ma3sim, echo=FALSE}
ma.sim = arima.sim(model=list(ma=c(-0.7, -0.5, 0.7)),n=500)

p1 = ma.sim  %>% autoplot() + theme_bw() +labs(y = 'MA(3): -.7, -.5, .7')
p2 = acf(ma.sim, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


### Example Plots of MA Processes [4]

```{r ma4sim, echo=FALSE}
ma.sim = arima.sim(model=list(ma=c(0.2, 0.4, 0.6, 0.8)),n=500)

p1 = ma.sim  %>% autoplot() + theme_bw() +labs(y = 'MA(4): .2, .4, .6, .8')
p2 = acf(ma.sim, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, ncol = 1, 
          labels = c('', ''), nrow = 2) 
```


# ARMA Models


### Mixed Autoregressive Moving Average Processes {.allowframebreaks}
Sometimes, if a really high order seems needed for an AR process, it may be better, instead, to add one or more MA term. This results in a mixed autoregressive moving average or an ARMA model.

**In general, an ARMA(p,q) model is given as**
\begin{equation*}
  y_t = \color{miamired}{\delta + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_p y_{t-p}} \color{blue}{+ \epsilon_t - \theta_1 \epsilon_{t-1}  - \dots - \theta_q\epsilon_{t-q} }
\end{equation*}

The ACF and PACF of the ARMA(p,q) process exhibit exponential decay exponential decay and/or damped sinusoid patterns. This makes
identification of the order of the ARMA(p,q) process difficult.

\begin{table}
  \centering
  \begin{tabular}{c P{2in} P{2in}}
\hline \hline
    \textbf{Model} & \textbf{ACF} & \textbf{PACF} \\ \hline
    AR(p) & Exponentially decays or damped sinusoidal pattern & Cuts off after lag $p$ \\ \hline
    MA(q) & Cuts off after lag $q$ & Exponentially decays or damped sinusoidal pattern \\ \hline 
    ARMA(p,q) & Exponentially decays or damped sinusoidal pattern & Exponentially decays or damped sinusoidal pattern \\ \hline \hline
  \end{tabular}
\end{table}




### Example Plot of an ARMA(1,1): AR=.6, MA=.8 [1]
```{r arma1sim1, echo=FALSE}
ar.sim1 = arima.sim(model=list(ar=c(.6),ma=c(0.8)),n=500)

ar.sim1 %>% autoplot() + theme_bw() +labs(y = 'ARMA')
```

### Example Plot of an ARMA(1,1): AR=.6, MA=.8 [2]
```{r arma1sim2, echo=FALSE}
acf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
```

### Example Plot of an ARMA(1,1): AR=.6, MA=.8 [3]
```{r arma1sim3, echo=FALSE}
pacf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
```

### Example Plot of an ARMA(1,1): AR=.6, MA=.8 [4]
```{r arma1sim4, echo=FALSE}
p1 = ar.sim1 %>% autoplot() + theme_bw() +labs(y = 'ARMA')
p2 = acf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
p3 = pacf(ar.sim1, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, p3, ncol = 1, 
          labels = c('ARMA(1,1): AR=.6, MA=.8', '', ''), nrow = 3) 
```


### Example Plots of an ARMA(1,1): AR=-.7, MA=-.6 [1]
```{r arma2sim1, echo=FALSE}
ar.sim2 = arima.sim(model=list(ar=c(-0.7),ma=c(-0.6)),n=500)

ar.sim2 %>% autoplot() + theme_bw() +labs(y = 'ARMA')
```


### Example Plots of an ARMA(1,1): AR=-.7, MA=-.6 [2]
```{r arma2sim2, echo=FALSE}
acf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
```


### Example Plots of an ARMA(1,1): AR=-.7, MA=-.6 [3]
```{r arma2sim3, echo=FALSE}
pacf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
```


### Example Plots of an ARMA(1,1): AR=-.7, MA=-.6 [4]
```{r arma2sim4, echo=FALSE}
p1 = ar.sim2 %>% autoplot() + theme_bw() +labs(y = 'ARMA')
p2 = acf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)
p3 = pacf(ar.sim2, plot = FALSE, lag.max = 25) %>% 
  autoplot() + theme_bw() + labs(title = NULL)

ggarrange(p1, p2, p3, ncol = 1, 
          labels = c('ARMA(1,1): AR=-.7, MA=-.6', '', ''), nrow = 3) 
```



# How to Fit an AR, MA, or ARMA Model

### The 5-Step Procedure {.allowframebreaks}

  1. Plot the data over time.  
  
  2. Do the data seem stationary? If necessary, conduct a test for stationarity.  
  
  3. Once you can assume stationarity, find the ACF plot.  
    a. If the ACF plot cuts off, fit an MA($q$), where $q=$ the cutoff point.  
    b. If the ACF plot dies down, find the PACF plot.  
      i. If the PACF plot cuts off, fit an AR($p$) model, where $p=$ the cutoff point.
      ii. If the PACF plot dies down, fit an ARMA ($p,q$) model.  
        You must iterate through $p$ and $q$ using a guess and check method starting with ARMA(1,1) models -- increment each by  1.
  
  4. Evaluate the model residuals and consider the ACF and PACF of the residuals.  
  
  5. If model fit is good, forecast future values.


**Note:** 

  - Often you will fit multiple models in Step 3 and compare models in Step 4 to select the best fit.



### Live Demo

Viscosity of a fluid is a measure that corresponds to “thickness”. For example, honey has a higher viscosity than water. A chemical company needs precise forecasts of the viscosity of a product in order to control product quality. Using the [18 - viscosity.csv](https://miamioh.instructure.com/courses/167327/files/23771321?module_item_id=3688201), we have 95 daily readings to use to develop a forecast. 

**In order to develop a forecast, let us first figure out what type of ARMA(p, q) model to fit and then develop the forecast.**


# Recap

### Summary of Main Points

\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Describe the behavior of the ACF and PACF of an AR(p) process.}
			\item \textbf{Describe the behavior of the ACF and PACF of an MA(q) process.}
			 \item \textbf{Describe the behavior of the ACF and PACF of an ARMA (p,q) process.}
			\item \textbf{Fit an ARMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARIMA model.}
	\end{itemize}
\end{block}


### Summary Table
\begin{table}
  \centering
  \begin{tabular}{c P{2in} P{2in}}
\hline \hline
    \textbf{Model} & \textbf{ACF} & \textbf{PACF} \\ \hline
    AR(p) & Exponentially decays or damped sinusoidal pattern & Cuts off after lag $p$ \\ \hline
    MA(q) & Cuts off after lag $q$ & Exponentially decays or damped sinusoidal pattern \\ \hline 
    ARMA(p,q) & Exponentially decays or damped sinusoidal pattern & Exponentially decays or damped sinusoidal pattern \\ \hline \hline
  \end{tabular}
\end{table}

 


---

\maketitle