---
title: "ISA 444: Business Forecasting"
subtitle: "21 - ARIMA Models"
author: Fadel M. Megahed
date: 'Spring 2022'
output:
  beamer_presentation:
    number_sections: false
    toc: false
    slide_level: 3
    latex_engine: xelatex
    includes: 
      in_header: structure.txt
classoption: "aspectratio=169"
always_allow_html: yes
bibliography: refs.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      verbose = FALSE,
                      progress = FALSE,
                      fig.align = "center",
                      fig.pos = 'p',
                      fig.width = 5.8,
                      fig.height= 2.9,
                      allowframebreaks = TRUE,
                      fig.margin=TRUE,
                      kable.force.latex = TRUE,
                      cache = TRUE)
options(kableExtra.latex.load_packages = FALSE)
pacman::p_load(kableExtra, tidyverse, tidyquant, xtable, magrittr, fpp2, ggpubr)
```

# Preface

### Recap of What we Have Covered Last Two Weeks

**ARMA Models:** Models we considered may have three components, an autoregressive component (AR), and a moving average component (MA).

\begin{block}{\textbf{Main Learning Outcomes from Last 2 Classes}}
  \begin{itemize}
      \item \textbf{Describe the behavior of the ACF and PACF of an AR(p) process.}
			\item \textbf{Describe the behavior of the ACF and PACF of an MA(q) process.}
			 \item \textbf{Describe the behavior of the ACF and PACF of an ARMA (p,q) process.}
			\item \textbf{Fit an ARMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARMA model.}
	\end{itemize}
\end{block}


### Summary Table for ARMA Models
\begin{table}
  \centering
  \begin{tabular}{c P{2in} P{2in}}
\hline \hline
    \textbf{Model} & \textbf{ACF} & \textbf{PACF} \\ \hline
    AR(p) & Exponentially decays or damped sinusoidal pattern & Cuts off after lag $p$ \\ \hline
    MA(q) & Cuts off after lag $q$ & Exponentially decays or damped sinusoidal pattern \\ \hline 
    ARMA(p,q) & Exponentially decays or damped sinusoidal pattern & Exponentially decays or damped sinusoidal pattern \\ \hline \hline
  \end{tabular}
\end{table}


### Finalizing Example from Class 18 and a Non-Graded Activity
Let us finalize the example in [18_class_code_full.R](https://miamioh.instructure.com/courses/167327/files/23786079?module_item_id=3689531).

After we go through the example, you will:
\begin{itemize}
\item fit an AR(2), check the residuals from the model, and examine the accuracy metrics from the fitted model
\item Based on the residuals and accuracy metrics determine whether the AR(2) fit is better than the MA(1) approach.
\end{itemize}




### Learning Outcomes for Today's Class
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Explain how ARIMA models work when compared to ARMA models.}
			\item \textbf{Fit an ARIMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARIMA model.}
			\item \textbf{Describe AIC, AICc, and BIC and how they are used to measure model fit.}
      \item \textbf{Describe the algorithm used within the auto.arima() function to fit an ARIMA model.}
      \item \textbf{Describe the results of the auto.arima() function.}
	\end{itemize}
\end{block}


# ARIMA Models


### ARIMA Models for Nonstationary Processes
When the time series is nonstationary, differencing can be used to transform the series. The ADF or KPSS tests can be used to test for stationarity, and the `ndiffs()` function is useful for determining the number of differences necessary to achieve stationarity.


### A Live-Example

In class, we will use the [GNP Data](https://fred.stlouisfed.org/series/GNP) to highlight how ARIMA models can be fit and used for forecasting.


### Non-graded In-Class Activity
You have 10 minutes to examine and model `series10` in [21 - InClass ARIMA Practice.csv](https://miamioh.instructure.com/courses/167327/files/23895210?module_item_id=3698033), where you should:  

  - Determine an appropriate model to fit based on examining the output from the `ndiffs()`, examining both the `acf()` and `pacf()` of your data (or differenced data), and the output from the `checkresiduals()`.
  
  - Once you fit an appropriate model, return its accuracy metrics and use it to forecast the values of the time-series for the next 10 time periods. **From your results, which of the following models did you pick? Why?**
    * ARIMA(1, 0, 1)  
    * ARIMA(2, 0, 1)  
    * ARIMA(1, 0, 2)  
    * ARIMA(2, 0, 2)

# Some Measures of Goodness of Fit

### Additional Measures of Goodness of Fit {.allowframebreaks}

**Akaike’s Information Criterion:** Akaike suggests measuring the goodness of fit of a model by balancing the error of the fit against the number of parameters in the model.\footnotemark

$AIC = 2k - 2\ln{\hat{(L)}}$, 

where $k$ is the number of parameters in the model and $\hat{L}$ is the sample likelihood function. The value of $k$ that gives a minimum AIC gives the best model. This is simply a penalty imposed on the error variance for the number of
parameters in the model.


**Bias Corrected AIC:** $AICc = AIC + \frac{2k^2 + 2k}{n - k - 1}$.

The AICc is usually preferred over the AIC.


**Bayesian Information Criterion:** $BIC = \ln{(n)}k - 2\ln{\hat{(L)}}$. 

BIC is also known as the Schwarz Information Criterion (SIC). The BIC has
a larger penalty for model size and tends to choose smaller models.


**Studies have shown:**  

  - BIC does well at getting correct model in large samples.  
  - AICc tends to get correct models in smaller samples with a large number of parameters.
  
  
**Why did we discuss these metrics today?**  

  - They were printed with some of the models that we have examined in class.   
  - They are used with the `auto.arima()`, which comes from the [forecast package](https://cran.r-project.org/web/packages/forecast/forecast.pdf) (loaded with `fpp2`).

\footnotetext{Slides are based on  \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}

# The `auto.arima()` Function

### The `auto.arima()` Function {.allowframebreaks}

The auto.arima() function can be used to automatically fit ARIMA models to a time series. It is a useful function, but it should be used with caution.\footnotemark

**The function**  

  - Uses “brute force” to fit many models and then selects the “best” based on a certain model criterion  
  - Works best when the data are stationary, but can be used with nonstationary data  
  - Tends to overfit the data  
  - Should always be used as a starting point for selecting a model and all models derived from the `auto.arima()` function should be properly vetted and evaluated.


The `auto.arima()` function combines  
  - Unit root tests (KPSS by default)  
  - Minimization of AICc
to obtain an $ARIMA(p,d,q)$ model using the following algorithm:


  1. Determine the number of differences, $d$, using a sequence of KPSS tests.  
  
  2. Determine $p$ and $q$ by minimizing AICc after differencing the data d times. Rather than considering all possible $p$ and $q$ combinations, a stepwise approach is taken.  
      - The best initial model with lowest AICc is selected from the following four:   
        - ARIMA (2,$d$,2),  
        - ARIMA (0,$d$,0),  
        - ARIMA (1,$d$,0), and   
        - ARIMA (0,$d$,1).  
        - *If d=0, then a constant, $c$, is included. If $d \ge 1$, then the constant is set to 0. The results of this step is called the current model.*  
      - Variations on the current model are considered by  
        - Vary $p$ and/or $q$ from current model by $\pm 1$  
        - Include/exclude $c$ from current model.  
        - The best model considered so far (either current or one of variations) becomes the *new current model*. 
      - Repeat previous step until no lower AICc can be found.
    
\footnotetext{Slides are based on  \href{https://miamioh.edu/fsb/directory/?up=/directory/farmerl2}{Dr. Allison Jones-Farmer's} lecture notes, Miami University, Spring 2020.}


### Live Coding: Example 1

In class, we will use a different snapshot of the GNP data that we have explored in class so far. The purpose of this different snapshot is two-fold:   

  - We are not sure whether the model we fit in the beginning of class would be appropriate (so this is somewhat of a revision of what to do in order to fit the "best" ARIMA model by hand).      
  
  - Walk you through the process of finding the model selected from the `auto.arima()`
  
```{r autoArima}
pacman::p_load(astsa) 
gnpData = gnp # will be loaded from the astsa package until 2002

# We will build on this example in class
```


### Live Coding: Example 2 (Rolling Origin)

We will discuss the example below in great detail in class.

\tiny
```{r manyModels, eval=FALSE}
btc_df = readRDS("btc_histDay.rds")
btc_df %>% ggplot(aes(x = date, y = close)) + geom_line()

nested_df = btc_df %>%
  select(date, close) %>% # keeping only relevant cols
  rsample::rolling_origin(initial = which(btc_df$date == "2020-12-31"),
                 assess =  1, # predict 1-day ahead
                 cumulative = TRUE) # each training sample grows by 1 day

nested_df %<>% mutate(
  data = map(.x = splits, .f = rsample::analysis),
  trainDates = map(.x = data, .f = `[[`, 'date' ),
  trainClose = map(.x = data, .f = extract2, 'close'),
  target_data = map(.x = splits, .f = rsample::assessment),
  target_date = map(.x = target_data, .f = extract2, 'date'),
  target_close = map_dbl(.x = target_data, .f= extract2, 'close')
) 

head(nested_df)

results_df = nested_df %>% 
  select(-c(data, trainDates, target_data, target_date))

results_df %<>%
  mutate(autoArima = map(.x = trainClose, .f = auto.arima) %>% 
           map(.f = forecast, h = 1) %>% map_dbl(.f = extract2, "mean") )
```

\normalsize



# Recap

### Summary of Main Points
\begin{block}{\textbf{Main Learning Outcomes}}
  \begin{itemize}
      \item \textbf{Explain how ARIMA models work when compared to ARMA models.}
			\item \textbf{Fit an ARMA model to a time series, evaluate the residuals of a fitted ARMA model to assess goodness of fit, use the Ljung-Box test for correlation among the residuals of an ARIMA model.}
			\item \textbf{Describe AIC, AICc, and BIC and how they are used to measure model fit.}
      \item \textbf{Describe the algorithm used within the auto.arima() function to fit an ARIMA model.}
      \item \textbf{Describe the results of the auto.arima() function.}
	\end{itemize}
\end{block}


### Things to Do to Prepare for Next Class

 
 - Go through the slides, examples and make sure you have a good understanding of what we have covered.  
 
 - Go through the posted assignment (see next slide).
 
 
### Graded Assignment 11: Evaluating your Understanding

Please go to \href{https://miamioh.instructure.com/courses/167327/quizzes/449743?module_item_id=3698031}{Canvas (click here)} and answer the questions. **The assignment is due on April 13, 2022 at 8:30 AM**

**What/Why/Prep?** The purpose of this assignment is to evaluate your understanding and retention of ARIMA modeling.

**General Guidelines:**  

\vspace{-0.5\baselineskip}

  - Individual assignment.  
  - This is **NOT** a timed assignment.    
  - Proctorio is NOT required for this assignment.  
  - You will need to have R installed (or accessible through the \href{https://virtualpc.fsb.miamioh.edu/RDWeb/Pages/en-US/default.aspx}{Remote Desktop})
 

---

\maketitle