---
title: "ISA 444: Business Forecasting"
subtitle: "16: PACF and AutoRegressive Moving Average Models" 
author: '<br>Fadel M. Megahed, PhD <br><br> Professor <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("twitter"), fill = "white")` @FadelMegahed](https://twitter.com/FadelMegahed) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2025"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["python"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:  
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE, 
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.75,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)


miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")
if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")
if(require(urbnmapr)==FALSE) devtools::install_github('UrbanInstitute/urbnmapr')
if(require(emo)==FALSE) devtools::install_github("hadley/emo")

pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               DataExplorer, scales, plotly, calendR, pdftools, # plots
               tmap, sf, urbnmapr, tigris, # maps
               bibliometrix, # for bibliometric analysis of my papers
               gifski, av, gganimate, ggtext, glue, extrafont, # for animations
               emojifont, emo, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE  
)

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "broadcast", "share_again", "search", "fit_screen", "editable", "clipboard"))

ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 12) + 
    ggplot2::theme(
      plot.title = ggtext::element_markdown(),
      plot.subtitle = ggtext::element_markdown(),
      plot.caption = ggtext::element_markdown(),
      legend.position = 'bottom',
      legend.title = ggtext::element_markdown(),
      legend.text = ggtext::element_markdown(),
      axis.title.x = ggtext::element_markdown(face = 'bold'),
      axis.title.y = ggtext::element_markdown(face = 'bold'),
      axis.text.x = ggtext::element_markdown(face= 'bold'),
      axis.text.y = ggtext::element_markdown(face = 'bold'),
      panel.grid.major = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
    )
  )

```

## Quick Refresher of Last Class

`r emo::ji("check")` Define a **stationary time series**.    

`r emo::ji("check")` Understand how plotting techniques can help identify non-stationarity.  

`r emo::ji("check")` Understand how we can **stabilize a non-stationary time series to achieve stationarity** (differencing to stabilize the mean, and log transformation to stabilize the variance).  

`r emo::ji("check")` Understand how to use **unit root tests** to test for stationarity.


---

## Learning Objectives for Today's Class

- Explain the **Partial Autocorrelation Function (PACF)**.  

- Explain the **Autoregressive (AR) model**, and how to identify the order of the AR model using the PACF.  

- Explain the **Moving Average (MA) model**, and how to identify the order of the MA model using the ACF.  

- Explain the **Autoregressive Moving Average (ARMA) model**.  

- Understand how to fit AR, MA, and ARMA models using the `ARIMA()` method from the [statsforecast](https://nixtlaverse.nixtla.io/statsforecast/docs/models/arima.html) Python package.


---
class: inverse, center, middle

# Revisting the Autocorrelation Function (ACF) and Examining the Partial Autocorrelation Function (PACF)

---

## Autocorrelation Function (ACF)  

- The autocorrelation is the correlation between a time series and a lagged version of itself.   

- The **ACF** is a function that **captures the autocorrelation of a time series at different lags**, which can .black[.bold[help in detecting stationarity and seasonality]].   

.font80[

$$\rho_k = \frac{Cov(Y_t, Y_{t-k})}{Var(Y_t)}$$

.left-code[
```{python acf1, eval=FALSE}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import acf
from statsmodels.graphics.tsaplots import plot_acf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['diff'], nlags=12, missing='drop')
  })
acf_df.head(2)

plot_acf(df['diff'], lags=12, missing='drop')
plt.title('ACF: First Difference in $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Autocorrelation')
```
]


.right-plot[
```{python acf_out1, echo=FALSE, out.width='100%', fig.align='center'}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import acf
from statsmodels.graphics.tsaplots import plot_acf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['diff'], nlags=12, missing='drop')
  })
acf_df.head(2)

plot_acf(df['diff'], lags=12, missing='drop')
plt.title('ACF: First Difference in $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Autocorrelation')
plt.show()
```
]
]


---
count: false

## Autocorrelation Function (ACF)  

- The autocorrelation is the correlation between a time series and a lagged version of itself.   

- The **ACF** is a function that **captures the autocorrelation of a time series at different lags**, which can .black[.bold[help in detecting stationarity and seasonality]].   

.font80[

$$\rho_k = \frac{Cov(Y_t, Y_{t-k})}{Var(Y_t)}$$

.left-code[
```{python acf2, eval=FALSE}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import acf
from statsmodels.graphics.tsaplots import plot_acf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['price'], nlags=12)
  })
acf_df.head(2)

plot_acf(df['price'], lags=12)
plt.title('ACF: $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Autocorrelation')
```
]


.right-plot[
```{python acf_out2, echo=FALSE, out.width='100%', fig.align='center'}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import acf
from statsmodels.graphics.tsaplots import plot_acf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['price'], nlags=12, missing='drop')
  })
acf_df.head(2)

plot_acf(df['price'], lags=12, missing='drop')
plt.title('ACF: $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Autocorrelation')
plt.show()
```
]
]

---
count: false
## Autocorrelation Function (ACF)  

- The autocorrelation is the correlation between a time series and a lagged version of itself.   

- The **ACF** is a function that **captures the autocorrelation of a time series at different lags**, which can .black[.bold[help in detecting stationarity and seasonality]].   

.font80[

$$\rho_k = \frac{Cov(Y_t, Y_{t-k})}{Var(Y_t)}$$

.left-code[
```{python acf3, eval=FALSE}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import acf
from statsmodels.graphics.tsaplots import plot_acf

df = pd.read_csv('../../data/aus_beer.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['beer'], nlags=12)
  })
acf_df.iloc[[1,4], : ] # subsetting lag 1 and 4

plot_acf(df['beer'], lags=12)
plt.title('ACF: Australian Beer Production')
plt.xlabel('Lag (Quarter)')
plt.ylabel('Autocorrelation')
```
]


.right-plot[
```{python acf_out3, echo=FALSE, out.width='100%', fig.align='center'}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import acf
from statsmodels.graphics.tsaplots import plot_acf

df = pd.read_csv('../../data/aus_beer.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['beer'], nlags=12)
  })

acf_df.iloc[[1,4], :]

plot_acf(df['beer'], lags=12)
plt.title('ACF: Australian Beer Production')
plt.xlabel('Lag (Quarter)')
plt.ylabel('Autocorrelation')
plt.show()
```
]
]


---

## Partial Autocorrelation Function (PACF)

- The **Partial Autocorrelation Function (PACF)** is a function that **captures the autocorrelation of a time series at different lags, while excluding the effect of intermediate lags**.  

.font80[
$$\phi_{k,k} = \frac{Cov(Y_t, Y_{t-k} | Y_{t-1}, Y_{t-2}, \ldots, Y_{t-k+1})}{Var(Y_t)}$$

.left-code[
```{python pacf1, eval=FALSE}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import pacf
from statsmodels.graphics.tsaplots import plot_pacf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['diff'].dropna(), nlags=12) #<<
  })
pacf_df.head(2)

plot_pacf(df['diff'].dropna(), lags=12) #<<
plt.title('PACF: First Difference in $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Partial Autocorrelation')
```
]

.right-plot[
```{python pacf_out1, echo=FALSE, out.width='100%', fig.align='center'}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import pacf
from statsmodels.graphics.tsaplots import plot_pacf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['diff'].dropna(), nlags=12) #<<
  })
pacf_df.head(2)

plot_pacf(df['diff'].dropna(), lags=12)
plt.title('PACF: First Difference in $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Partial Autocorrelation')
plt.show()

```
]
]


---
count: false

## Partial Autocorrelation Function (PACF)

- The **Partial Autocorrelation Function (PACF)** is a function that **captures the autocorrelation of a time series at different lags, while excluding the effect of intermediate lags**.  

.font80[
$$\phi_{k,k} = \frac{Cov(Y_t, Y_{t-k} | Y_{t-1}, Y_{t-2}, \ldots, Y_{t-k+1})}{Var(Y_t)}$$

.left-code[
```{python pacf2, eval=FALSE}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import pacf
from statsmodels.graphics.tsaplots import plot_pacf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['price'], nlags=12)
  })
pacf_df.head(2)

plot_pacf(df['price'], lags=12)
plt.title('PACF: $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Partial Autocorrelation')
```
]

.right-plot[
```{python pacf_out2, echo=FALSE, out.width='100%', fig.align='center'}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import pacf
from statsmodels.graphics.tsaplots import plot_pacf

df = pd.read_csv('../../data/kroger_stock_2024.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['price'], nlags=12)
  })
pacf_df.head(2)

plot_pacf(df['price'], lags=12)
plt.title('PACF: $KR Prices')
plt.xlabel('Lag (Day)')
plt.ylabel('Partial Autocorrelation')
plt.show()

```
]
]


---
count: false

## Partial Autocorrelation Function (PACF)

- The **Partial Autocorrelation Function (PACF)** is a function that **captures the autocorrelation of a time series at different lags, while excluding the effect of intermediate lags**.

.font80[
$$\phi_{k,k} = \frac{Cov(Y_t, Y_{t-k} | Y_{t-1}, Y_{t-2}, \ldots, Y_{t-k+1})}{Var(Y_t)}$$

.left-code[
```{python pacf3, eval=FALSE}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import pacf
from statsmodels.graphics.tsaplots import plot_pacf

df = pd.read_csv('../../data/aus_beer.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['beer'], nlags=12)
  })
pacf_df.iloc[[1,4], : ] # subsetting lag 1 and 4

plot_pacf(df['beer'], lags=12)
plt.title('PACF: Australian Beer Production')
plt.xlabel('Lag (Quarter)')
plt.ylabel('Partial Autocorrelation')
```
]

.right-plot[
```{python pacf_out3, echo=FALSE, out.width='100%', fig.align='center'}
import matplotlib.pyplot as plt
import pandas as pd
from statsmodels.tsa.stattools import pacf
from statsmodels.graphics.tsaplots import plot_pacf

df = pd.read_csv('../../data/aus_beer.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['beer'], nlags=12)
  })
pacf_df.iloc[[1,4], : ] # subsetting lag 1 and 4

plot_pacf(df['beer'], lags=12)
plt.title('PACF: Australian Beer Production')
plt.xlabel('Lag (Quarter)')
plt.ylabel('Partial Autocorrelation')
plt.show()
```
]
]


---

## Comments on the PACF Computation in `statsmodels`

- `statsmodels` uses the **Yule-Walker equations, with sample-size adjustment,** by default. 

```{r statsmodels_pacf, echo=FALSE}
knitr::include_url("https://www.statsmodels.org/stable/generated/statsmodels.tsa.stattools.pacf.html", height = '450px')
```

---

## Comments on the PACF Computation in `statsmodels`

- The use of the Yule-Walker (`yw`) equations in `statsmodels` is a **common approach** to computing the PACF.   

- The **sample-size adjustment** in `statsmodels` is used to **correct for the bias** in the sample autocorrelation function estimates. 

- The **default** in `statsmodels` is to use the **sample-size adjustment**, hence `yw` is sample-size adjusted.

-  Levinson-Durbin (`ld`) showed that the Yule-Walker equations can be rewritten as a set of linear equations that can be solved using a **recursive** algorithm. Hence,  
  + `ld` is a .black[.bold[computationally efficient implementation]] of `yw`, and  
  + `ldb` is .black[.bold[computationally efficient implementation]] of `ywm`. 

- **On the other hand**, the `pacf()` in `r fontawesome::fa("r-project", fill=miamired)`  (`stats` package) uses the **biased implementation** the Yule-Walker as its **default**, i.e.,  `r fontawesome::fa("r-project", fill=miamired)` is equivalent to `ywm` or `ldb` in `statsmodels`.


---

## Verifying PACF Computations in `statsmodels` & `r fontawesome::fa("r-project", fill=miamired)`

.left-code[
.font80[
```{python pacf_statsmodels, eval=FALSE}
import numpy as np
import pandas as pd
from statsmodels.tsa.stattools import pacf

# generate a random time series
np.random.seed(2025)
ts = np.random.normal(size=100, loc=50, scale=10)  

# computing the PACF using different methods
pacf_df = pd.DataFrame({
    'lag': range(13),
    'yw': pacf(ts, nlags=12, method='yw'),
    'ld': pacf(ts, nlags=12, method='ld'),
    'ywm': pacf(ts, nlags=12, method='ywm'),
    'ldb': pacf(ts, nlags=12, method='ldb')
})

# print the results
pacf_df

# save the original ts to reuse in R
pd.DataFrame({'ts': ts}).to_csv(
  '../../data/sim_ts.csv', index=False)

# save the PACF results to reuse in R
pacf_df.to_csv(
  '../../data/pacf_results.csv', index=False)
```
]
]

.right-plot[
.font80[
```{python pacf_statsmodels_out, echo=FALSE, out.width='100%', fig.align='center', ref.label='pacf_statsmodels'}

```
]
]


---

## Verifying PACF Computations in `statsmodels` & `r fontawesome::fa("r-project", fill=miamired)`

### `r fontawesome::fa("r-project", fill=miamired)` Computations and Combining Results

```{r pacf_r, eval=FALSE}
ts = read.csv('../../data/sim_ts.csv')
pacf_results = read.csv('../../data/pacf_results.csv')

# r uses biased implementation of Yule-Walker and does not include lag 0
pacf_r = pacf(ts$ts, lag.max=12, plot=FALSE)
pacf_results$r = c(0, pacf_r[[1]]) # prefix w/ lag 0

pacf_results[2:6,]
```

### The Combined Results from `statsmodels` and `r fontawesome::fa("r-project", fill=miamired)`

```{r pacf_r_out, echo=FALSE, out.width='100%', fig.align='center', paged.print=TRUE, cols.print='lag:yw:ld:ywm:ldb:r', rownames.print=FALSE}
formals(print.data.frame)$row.names = FALSE

ts = read.csv('../../data/sim_ts.csv')
pacf_results = read.csv('../../data/pacf_results.csv')

# r uses biased implementation of Yule-Walker and does not include lag 0
pacf_r = pacf(ts$ts, lag.max=12, plot=FALSE)
pacf_results$r = c(0, pacf_r[[1]]) # prefix w/ lag 0

print(pacf_results[2:6,])

```

---

## Assumptions of the Yule-Walker Equations

For using the Yule-Walker equations to compute the PACF, the following assumptions are made:

- The time series is **stationary**.  

- The time series is in type **AR(p)** form, but the **order p is unknown**.  

- Having the prediction $p$, we use the appropriate **AR(p)** model to estimate the coefficients (i.e., the obtained correlation coefficients are preliminary).  
- We will find the estimators for $\phi_1, \phi_2, \ldots, \phi_p$ later using the maximum likelihood method via fitting the AR(p) model.

.footnote[
<html>
<hr>
</html>

**Source:** Łukasz Balbus (2020). Lecture Notes on Econometrics. University of Zielona Góra, Zielona Góra, Poland. Available at: <https://staff.uz.zgora.pl/lbalbus/Zdalne%20EDiF%2010.11.2020.pdf>.
]

---

## ACF Patterns to Look for: Dying Down

```{r arima_simulations, include=FALSE}
# simulate an AR(1) process
set.seed(2025)

ts1 = arima.sim(n=200, list(ar=0.5))
ts2 = arima.sim(n=200, list(ar=-0.5))
ts3 = arima.sim(n=200, list(ar=c(-0.5, 0.3)))

ts4 = arima.sim(n=200, list(ma=0.5))
ts5 = arima.sim(n=200, list(ma=-0.5))
ts6 = arima.sim(n=200, list(ma=c(-0.5, 0.3)))

ts7 = arima.sim(n=200, list(ar=0.5, ma=0.5))
ts8 = arima.sim(n=200, list(ar=-0.5, ma=-0.5))
ts9 = arima.sim(n=200, list(ar=c(-0.5, 0.3), ma=c(-0.5, 0.3)))

df = data.frame(ts1, ts2, ts3, ts4, ts5, ts6, ts7, ts8, ts9)

readr::write_csv(df, '../../data/simulated_ts.csv')

```

- **Dying Down**: The ACF will **exponentially decay to zero relatively quickly** $\longrightarrow$ this is a sign of a possible **AR** Data Generating Process (DGP).

```{python dying_down, echo=FALSE, results='hide', fig.show='first'}
import sys
import os 
utils_path = os.path.abspath(os.path.join(os.path.dirname(__name__), '..', '..'))
sys.path.append(utils_path)

from utils.plotting import plot_acf_colors

ts = pd.read_csv('../../data/simulated_ts.csv')['ts1']

acf_vals = acf(ts, nlags=20, fft=False)

# plot the ACF
f, ax = plt.subplots(figsize=(10, 4))
plot_acf(ts, lags=20, ax=ax)
ax = plot_acf_colors(ax)
plt.title('ACF: Dying Down', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')

plt.plot(acf_vals, color='darkgray', linestyle='dashed', lw=0.5)
plt.show()
```

---

## ACF Patterns to Look for: Dying Down (Sinsoidal)

- **Dying Down (Sinsoidal)**: The ACF will **exponentially decay to zero with some oscillations** $\longrightarrow$ a possible **AR** DGP, with at least one negative correlation coefficient.

```{python dying_down_sinsoidal1, echo=FALSE, results='hide', fig.show='first'}
ts = pd.read_csv('../../data/simulated_ts.csv')['ts2']

acf_vals = acf(ts, nlags=20, fft=False)

# plot the ACF
f, ax = plt.subplots(figsize=(10, 4))
plot_acf(ts, lags=20, ax=ax)
ax = plot_acf_colors(ax)
plt.title('ACF: Dying Down (Sinsoidal)', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')

# Overlay the sinsioidal pattern
plt.plot(acf_vals, color='darkgray', linestyle='dashed', lw=0.5)
plt.show()
```

---
count: false

## ACF Patterns to Look for: Dying Down (Sinsoidal)

- **Dying Down (Sinsoidal)**: The ACF will **exponentially decay to zero with some oscillations** $\longrightarrow$ a possible **AR** DGP, with at least one negative correlation coefficient.

```{python dying_down_sinsoidal2, echo=FALSE, results='hide', fig.show='first'}
ts = pd.read_csv('../../data/simulated_ts.csv')['ts3']

acf_vals = acf(ts, nlags=20, fft=False)

# plot the ACF
f, ax = plt.subplots(figsize=(10, 4))
plot_acf(ts, lags=20, ax=ax)
ax = plot_acf_colors(ax)
plt.title('ACF: Dying Down (Sinsoidal)', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')

# Overlay the sinsioidal pattern
plt.plot(acf_vals, color='darkgray', linestyle='dashed', lw=0.5)
plt.show()
```


---

## ACF Patterns to Look for: Cutting Off

- **Cutting Off**: The ACF will **cut off after a certain lag** $\longrightarrow$ a possible **MA** DGP.


```{python cutting_off1, echo=FALSE, results='hide', fig.show='first'}
ts = pd.read_csv('../../data/simulated_ts.csv')['ts4']

acf_vals = acf(ts, nlags=20, fft=False)

# plot the ACF
f, ax = plt.subplots(figsize=(10, 4))
plot_acf(ts, lags=20, ax=ax)
ax = plot_acf_colors(ax)
plt.title('ACF: Cutting Off at Lag 1', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')

plt.axvline(x=1.5, color='black', linestyle='--', lw=0.75)
plt.annotate(
  'Correlations after cutoff\n (Not significant, i.e., not\n  statistically different from 0)', 
  xy=(1.5, 0.3), xytext=(3, 0.3), fontsize=10, fontweight='bold'
  )
plt.annotate(
  '', xy=(2.85, 0.35), xytext=(1.53, 0.35), 
  arrowprops=dict(facecolor='#C3142D', shrink=0.05, edgecolor='#C3142D'),
  fontsize=10, fontweight='bold'
  )

plt.plot(acf_vals, color='darkgray', linestyle='dashed', lw=0.5)
plt.show()
```

---
count: false

## ACF Patterns to Look for: Cutting Off

- **Cutting Off**: The ACF will **cut off after a certain lag** $\longrightarrow$ a possible **MA** DGP.


```{python cutting_off2, echo=FALSE, results='hide', fig.show='first'}
ts = pd.read_csv('../../data/simulated_ts.csv')['ts5']

acf_vals = acf(ts, nlags=20, fft=False)

# plot the ACF
f, ax = plt.subplots(figsize=(10, 4))
plot_acf(ts, lags=20, ax=ax)
ax = plot_acf_colors(ax)
plt.title('ACF: Cutting Off at Lag 1', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')

plt.axvline(x=1.5, color='black', linestyle='--', lw=0.75)
plt.annotate(
  'Correlations after cutoff\n (Not significant, i.e., not\n  statistically different from 0)', 
  xy=(1.5, 0.3), xytext=(3, 0.3), fontsize=10, fontweight='bold'
  )
plt.annotate(
  '', xy=(2.85, 0.35), xytext=(1.53, 0.35), 
  arrowprops=dict(facecolor='#C3142D', shrink=0.05, edgecolor='#C3142D'),
  fontsize=10, fontweight='bold'
  )

plt.plot(acf_vals, color='darkgray', linestyle='dashed', lw=0.5)
plt.show()
```

---
count: false

## ACF Patterns to Look for: Cutting Off

- **Cutting Off**: The ACF will **cut off after a certain lag** $\longrightarrow$ a possible **MA** DGP.


```{python cutting_off3, echo=FALSE, results='hide', fig.show='first'}
ts = pd.read_csv('../../data/simulated_ts.csv')['ts6']

acf_vals = acf(ts, nlags=20, fft=False)

# plot the ACF
f, ax = plt.subplots(figsize=(10, 4))
plot_acf(ts, lags=20, ax=ax)
ax = plot_acf_colors(ax)
plt.title('ACF: Cutting Off at Lag 2', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')

plt.axvline(x=2.5, color='black', linestyle='--', lw=0.75)
plt.annotate(
  'Correlations after cutoff\n (Not significant, i.e., not\n  statistically different from 0)', 
  xy=(2.5, 0.3), xytext=(4, 0.3), fontsize=10, fontweight='bold'
  )
plt.annotate(
  '', xy=(3.85, 0.35), xytext=(2.53, 0.35), 
  arrowprops=dict(facecolor='#C3142D', shrink=0.05, edgecolor='#C3142D'),
  fontsize=10, fontweight='bold'
  )


plt.plot(acf_vals, color='darkgray', linestyle='dashed', lw=0.5)
plt.show()
```


---

## ACF Patterns and Data Generating Processes

.left-code[

.white[ABC]

- **White Noise:** ACF will .black[.bold[show no pattern.]]

<br><br><br>

- **Random Walk:** ACF will .black[.bold[slowly decay.]]  

<br><br>

- **Seasonal Patterns:** ACF will .black[.bold[show a pattern at regular intervals.]] 
]

.right-plot[
```{python other_acf_patterns, echo=FALSE, results='hide', out.width='100%'}
df = pd.read_csv('../../data/kroger_stock_2024.csv')

fig, axes = plt.subplots(nrows=3, ncols=1, figsize=(5, 5) )
acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['diff'], nlags=12, missing='drop')
  })
acf_df.head(2)

plot_acf(df['diff'], lags=12, missing='drop', ax=axes[0])
axes[0] = plot_acf_colors(axes[0])
axes[0].set_title('ACF: First Difference in $KR Prices (White Noise)')
axes[0].set_ylabel('Autocorrelation')


acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['price'], nlags=12)
  })
acf_df.head(2)

plot_acf(df['price'], lags=12, ax=axes[1])
axes[1] = plot_acf_colors(axes[1])
axes[1].set_title('ACF: $KR Prices (Non-Stationarity [Random Walk])')
axes[1].set_ylabel('Autocorrelation')


df = pd.read_csv('../../data/aus_beer.csv')

acf_df = pd.DataFrame({
    'lag': range(0, 13),
    'acf': acf(df['beer'], nlags=12)
  })
acf_df.iloc[[1,4], : ] # subsetting lag 1 and 4

plot_acf(df['beer'], lags=12, ax=axes[2])
axes[2] = plot_acf_colors(axes[2])
axes[2].set_title('ACF: Beer Prod (Non-Stationarity [Seasonality])')
axes[2].set_ylabel('Autocorrelation')

plt.subplots_adjust(hspace=0.5) # Increase vertical spacing
plt.tight_layout()
plt.show()
```
]


---

## PACF Patterns to Look For: Cutting Off

- If the **PACF cuts off after the ACF has decayed down**, this is indicates that an .black[.bold[AR DGP is suitable]]. Let us consider the PACF of Kroger's stock price.

```{python pacf_cutting_off, echo=FALSE, results='hide', fig.show='first'}
df = pd.read_csv('../../data/kroger_stock_2024.csv')

pacf_df = pd.DataFrame({
    'lag': range(0, 13),
    'pacf': pacf(df['price'], nlags=12, method ='ywm')
  })

fig, ax = plt.subplots(figsize=(10, 4))
plot_pacf(df['price'], lags=12, ax=ax)
ax = plot_acf_colors(ax)
plt.title('PACF: $KR Prices', fontweight='bold')
plt.xlabel('Lag (Day)', fontweight='bold')
plt.ylabel('Partial Autocorrelation', fontweight='bold')

plt.axvline(x=1.5, color='black', linestyle='--', lw=0.75)
plt.annotate(
  'Correlations after cutoff\n (Not significant, i.e., not\n  statistically different from 0)', 
  xy=(1.5, 0.3), xytext=(3, 0.3), fontsize=10, fontweight='bold'
  )
plt.annotate(
  '', xy=(2.85, 0.35), xytext=(1.53, 0.35), 
  arrowprops=dict(facecolor='#C3142D', shrink=0.05, edgecolor='#C3142D'),
  fontsize=10, fontweight='bold'
  )

plt.show()
  
```

---
class: inverse, center, middle

# Autoregressive (AR) Models

---

## AR Model

In an autoregressive model, we forecast the variable of interest using a linear combination of past values of the variable. The term **autoregressive** indicates that it is a regression of the variable against itself.  Thus, an autoregressive model of order $p$, denoted as AR(p), can be written as:

$$y_t = c + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_p y_{t-p} + \epsilon_t,$$
where $\epsilon_t$ is white noise. 


**Note:** The AR model **looks similar** to a linear regression model that uses the dependent relationship between an observation and its lagged observations. The **difference** .black[.bold[lies in how their coefficients are estimated in practice.]]  
  + AR models $\longrightarrow$ .black[.bold[often]] utilize MLE while taking into account the time series structure.   
  + Linear regression $\longrightarrow$  uses ordinary least squares (OLS).  
  

---

## AR (MLE) vs. Linear Regression (OLS)

Since you used `r fontawesome::fa("r-project", fill=miamired)` for linear regression, let's illustrate the differences between AR and linear regression using base `r fontawesome::fa("r-project", fill=miamired)`.  

Let us simulate the following AR(1) process: $y_t = -0.2 + 0.7y_{t-1} + \epsilon_t$, where $\epsilon_t \sim N(0, 1)$.  

.left-code[
.font80[
```{r ar1_simulation, eval=FALSE}
set.seed(2025)

# Simulate y[t] = -0.2 + 0.7 * y[t-1] + error,
n = 200
y = 0
phi = 0.7
intercept = -0.2
errors = rnorm(n, mean = 0, sd = 1)

for (t in 2:n) {
  y[t] = intercept + phi*y[t - 1] + errors[t]
}

plot(
  y, type = "l", lwd = 2, xlab = "Time", 
  ylab = "y[t]", main = "AR(1) Simulation"
  )

write.csv(
  data.frame(y), '../../data/ar1_simulation.csv'
  )
```
]
]

.right-plot[
```{r ar1_simulation_out, echo=FALSE, fig.width=5, fig.height=3, ref.label='ar1_simulation'}

```
]

---

## Simulation: Examining the ACF and PACF

Let us examine the ACF and PACF of our simulated data. 

.left-code[
```{python acf_simulated_ar1, echo=FALSE}
import pandas as pd
import matplotlib.pyplot as plt
from statsmodels.tsa.stattools import acf, pacf
from statsmodels.graphics.tsaplots import plot_acf, plot_pacf

df = pd.read_csv('../../data/ar1_simulation.csv')

fig, axes = plt.subplots(figsize=(5, 4.25))
plot_acf(df['y'], lags=20, ax = axes)
axes = plot_acf_colors(axes)

plt.title('ACF: AR(1) Simulation', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')
plt.show()
```
]

.right-plot[
```{python pacf_simulated_ar1, echo=FALSE}
import pandas as pd
import matplotlib.pyplot as plt
from statsmodels.tsa.stattools import acf, pacf
from statsmodels.graphics.tsaplots import plot_acf, plot_pacf

df = pd.read_csv('../../data/ar1_simulation.csv')

fig, axes = plt.subplots(figsize=(5, 4.25))
plot_pacf(df['y'], lags=20, ax=axes)
axes = plot_acf_colors(axes)

plt.title('PACF: AR(1) Simulation', fontweight='bold')
plt.xlabel('Lag', fontweight='bold')
plt.ylabel('Autocorrelation', fontweight='bold')
plt.show()

```
]


---

## Simulation: The Linear Regression Model (Per ISA291)

.left-code[
.font80[
```{r linear_regression, eval=FALSE, results='asis'}
df = data.frame(y = y)
df$lag1 = dplyr::lag(df$y, 1)
df_reg = na.omit(df)

lm_model = lm(y ~ lag1, data = df_reg)
stargazer::stargazer(lm_model, type = 'html')
```

```{r linear_regression_out, echo=FALSE, fig.width=5, fig.height=3.6, results='hide', fig.show='asis'}
# Create a data frame with lagged values
df = data.frame(y = y)
df$lag1 = dplyr::lag(df$y, 1)

# Remove rows with NA values
df_reg = na.omit(df)

# Fit the OLS regression model using lm()
lm_model = lm(y ~ lag1, data = df_reg)

# diagnostic plots
forecast::checkresiduals(lm_model)
```

]
]

.right-plot[
.font80[
```{r linear_regression_out2, echo=FALSE, ref.label='linear_regression', results='asis'}

```
]
]


---

## Simulation: The AR(1) Model Results

.left-code[
.font80[
```{python ar1_model, eval=FALSE}
from statsforecast.models import ARIMA
from statsforecast import StatsForecast
df = (
  pd.read_csv('../../data/ar1_simulation.csv')
  .assign(
    unique_id = 'simulated_data',
    ds = pd.date_range(start='2024-01-01', periods=200, freq='D')
  )
)

# typical statforecast workflow
models = [ARIMA(order=(1,0,0))]
sf = StatsForecast(models=models, freq='D')
sf.fit(df)

# getting the model coefficients
# first 0 is for unique_id (only 1 unique_id) and
# second 0 is for the ARIMA model (only 1 model)
phi1 = sf.fitted_[0][0].model_.get('coef')['ar1']
intercept = sf.fitted_[0][0].model_.get('coef')['intercept']

print(
  'The AR(1) model equation is:\n',
  'y[t] = {:.3f} + {:.3f} * y[t-1] + e[t]'
  .format(intercept, phi1)
  )
```
]
]

.right-plot[
.font80[
```{python ar1_model_out, echo=FALSE, ref.label='ar1_model', results='hold'}

```

<br><br><br>
.center[
.black[
.bold[The AR(1) and the linear regression model are quite similar, but are not identical in this case. The differences arise from the estimation methods used in each model.]
]
]
]
]


---

## Determining the Order of the AR Model

- We can usually tell from the ACF that there is an autoregressive (AR) component to the data because the ACF plot tends to geometrically decrease in magnitude (i.e., "die down").

- The **order** of an AR Process refers to how many lags you include in the autoregressive model.

- Because the ACF of the AR model is a mixture, the **ACF is not useful for determining the order of the AR process.**

- Thus, the .black[.bold[ACF helps us to know that we have an AR model, but not which AR model to fit!]]

<br>

- The **Partial Autocorrelation Function (PACF)** is used to determine the order of the AR model.  
  + For an AR(p) model, the PACF between $y_t$ and $y_{t-k}$ should be 0 $\forall \ k > p$.  
  + The PACF will cut off after lag $p$.

---
class: inverse, center, middle

# Moving Average (MA) Models


---

## The Moving Average (MA) Model

Rather than using past values of the forecast variable in a regression, a moving average model uses **past forecast errors in a regression-like model**,

$$y_t = c + \epsilon_t + \theta_1 \epsilon_{t-1} + \theta_2 \epsilon_{t-2} + \dots + \theta_q \epsilon_{t-q},$$
where:  

- $\epsilon_t$ is white noise; $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$.
- Unlike AR models, **past values of $y_t$ do not appear**; instead, $y_t$ depends on past errors.  
- The errors $\epsilon_t$ **are unobservable**, meaning they **must be inferred from the data**.


---

## How do we Estimate $\theta$ if $\epsilon_t$ is Unobserved?

**A. Maximum Likelihood Estimation (MLE)**

MLE estimates the parameters by maximizing the likelihood function assuming the residuals $\epsilon_t$ follow a normal distribution. The process is:  

1. Assume an initial set of MA parameters $\theta_1, \theta_2, \ldots, \theta_q$.  

2. Use an **interative algorithm* (e.g., Expectation-Maximization, Kalman Filtering, or numerical optimization)** to estimate the most likely error terms $\epsilon_t$.  

3. Maximize the likelihood of the observed data given those estimated errors.  



**B. Nonlinear Least Squares (NLS)**

1. Instead of directly computing errors, we **minimize the sum of squared residuals using a nonlinear function of parameters.**  

2. This requires **iterative numerical techniques** (e.g., gradient descent or Newton-Raphson).


---

## Forecasting with MA Model if $\epsilon_t$ Is Unobserved?

### 1. One-Step-Ahead Forecast \(t+1\):

- Since $\epsilon_t$ is unknown in the future, we assume $\mathbb{E}[\epsilon_{t+h}] = 0$.  
- The forecast equation uses the most recent residuals $\epsilon_t, \epsilon_{t-1}, \ldots$:  
$$\hat{y}_{t+1} = c + \hat{\theta}_1 \hat{\epsilon}_t + \hat{\theta}_2 \hat{\epsilon}_{t-1} + \dots + \hat{\theta}_q \hat{\epsilon}_{t+1-q}$$

### 2. Multi-step Forecasts \(t+h\)  

- For **longer horizons**, all future errors are assumed to be zero (i.e., $\mathbb{E}[\epsilon_{t+h}] = 0$ for $h \ge 1$).  

- This causes the forecast to **revert to the mean** \( c \) over time, as past shocks/errors fade out.


---

## Determining the Order of the MA Model  


- The **order** of an MA Process refers to how many lags you include in the moving average model.  

- The **ACF** of an MA process will show a **sharp cut-off** after lag $q$ (the order of the MA model).  

- This cutting off is theoretically related to the **population autocorrelation function** of the MA(q) process at lag $k$:  

$$\rho(k) = 
  \begin{cases}
    \frac{(\theta_k + \theta_1 \theta_{k+1} + \dots + \theta_{q-k} \theta_{q})}{1+ \theta_1^2 + \dots + \theta_q^2}, & k = 1, \, 2, \dots, \, q \\
    0, & k > q
  \end{cases}.$$

- While the equation looks quite complex, the **key takeaway** is that the ACF is not significant (will cut off after lag $q$) for an MA(q) process.


---
class: inverse, center, middle

# Autoregressive Moving Average (ARMA) Models


---

## Autoregressive Moving Average (ARMA) Models

Sometimes, if a really high order seems needed for an AR process, it may be better to add one or more MA term. This results in a mixed autoregressive moving average (ARMA) model.

**In general, an ARMA(p,q) model is given as** $$\underbrace{y_t}_{\substack{\text{Predicted}\\ \text{series}}} = \color{darkgrey}{\underbrace{\delta}_{constant}} + \color{#C3142D}{\underbrace{\phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_p y_{t-p}}_{\substack{\text{Linear combination of Lags of Y}\\ \text{up to }p \text{ lags}}}} + \color{darkorange}{\underbrace{\epsilon_t + \theta_1 \epsilon_{t-1}  + \dots + \theta_q\epsilon_{t-q}}_{\substack{\text{Linear combination of Lagged forecast}\\ \text{errors up to }q \text{ lags}}}}$$


.font90[
The **ACF and PACF of the ARMA(p,q) process exhibit exponential decay exponential decay/ damped sinusoidal patterns**. This makes identifying the order of the ARMA(p,q) difficult.
]

.font80[

<html>
<table>
<tr>
<td><b>Model</b></td>
<td><b>ACF</b></td>
<td><b>PACF</b></td>
</tr>
<tr>
<td><b>AR(p)</b></td>
<td>Exponentially decays or <br>
damped sinusoidal pattern</td>
<td> Cuts off after lag p </td>
</tr>
<tr>
<td><b>MA(q)</b></td>
<td> Cuts off after lag q </td>
<td>Exponentially decays or <br>
damped sinusoidal pattern</td>
</tr>
<tr>
<td><b>ARMA(p,q)</b></td>
<td>Exponentially decays or <br>
damped sinusoidal pattern</td>
<td>Exponentially decays or <br>
damped sinusoidal pattern</td>
</tr>
</table>
</html>
]


---

## Fitting AR, MA, and ARMA Models

.font90[
1. Plot the data over time.  

2. Do the data seem stationary? If necessary, conduct a test for stationarity.  

3. Once you can assume stationarity, find the ACF plot.  
  - If the ACF plot cuts off, fit an MA(q), where $q=$ the cutoff point.  
  - If the ACF plot dies down, find the PACF plot.  
    + If the PACF plot cuts off, fit an AR(p) model, where $p=$ the cutoff point.
    + If the PACF plot dies down, fit an ARMA (p,q) model.  
      * You must iterate through $p$ and $q$ using a guess and check method starting with ARMA(1,1) models -- increment each by  1.

4. Evaluate the model residuals and consider the ACF and PACF of the residuals.  

5. If model fit is good, forecast future values.
]

.font90[
**Note:** Often you will fit multiple models in Step 3 and compare models in Step 4 to select the best fit.
]


---

## Fitting AR, MA, and ARMA Models in Python

Let us examine the [viscosity data](https://raw.githubusercontent.com/fmegahed/isa444/refs/heads/main/data/viscosity.csv). For this demo, we will only fit ARMA models on the **entire series**, and examine the residuals to see the goodness of fit.

However, it should be noted that **you are also expected to utilize the cross-validation method to evaluate the model's predictive performance (which we do not do in this demo to save time).**

```{python viscosity_data, eval=FALSE, include = FALSE}
import requests

req = requests.get('https://raw.githubusercontent.com/fmegahed/isa444/refs/heads/main/utils/fitted.py')
exec(req.text)

# !pip install utilsforecast statsforecast
import pandas as pd
import matplotlib.pyplot as plt

from utilsforecast.plotting import plot_series
from coreforecast.differences import num_diffs
from statsmodels.graphics.tsaplots import plot_acf, plot_pacf

from statsforecast import StatsForecast
from statsforecast.models import AutoARIMA, ARIMA

import statsmodels.api as sm

df = (
  pd.read_csv('https://raw.githubusercontent.com/fmegahed/isa444/refs/heads/main/data/viscosity.csv')
  )

df = (
  df
  .assign(
    ds = pd.date_range(start='2024-01-01', periods=len(df), freq='D'),
    unique_id = 'viscosity'
    )
  [['unique_id', 'ds', 'y']]
)

display( plot_series(df) )

# Does this need differencing
print('\n Our series need ', num_diffs(df['y']), 'differences to be stationary')

fig, axes = plt.subplots(1, 2, figsize=(15, 5))
axes[0].set_title('ACF')
axes[1].set_title('PACF')
plot_acf(df['y'], lags=20, ax=axes[0])
plot_pacf(df['y'], lags=20, ax=axes[1])
plt.show()

print('Both the ACF and PACF exhibit a damped sinsodial pattern; hence we should be fitting an ARMA type model')

models = [
    ARIMA(order=(1, 0, 1), alias = 'arma11'),
    ARIMA(order=(2, 0, 1), alias = 'arma21'),
    ARIMA(order=(1, 0, 2), alias = 'arma12'),
    ARIMA(order=(2, 0, 2), alias = 'arma22')
]

sf = StatsForecast(models = models, freq='D')
fitted = sf.fit(df)

res = create_res_df(df, fitted= fitted, model_names=['arma11', 'arma21', 'arma12', 'arma22'])

lb_results = pd.concat(
    [sm.stats.acorr_ljungbox(res.arma11, lags=[10], model_df=2),
     sm.stats.acorr_ljungbox(res.arma21, lags=[10], model_df=3),
     sm.stats.acorr_ljungbox(res.arma12, lags=[10], model_df=3),
     sm.stats.acorr_ljungbox(res.arma22, lags=[10], model_df=4)]
)
lb_results.set_axis(['arma11', 'arma21', 'arma12', 'arma22'])

```




---

class: inverse, center, middle

# Recap

---

## Summary of Main Points

By now, you should be able to do the following:  

- Explain the **Autoregressive (AR) model**, and how to identify the order of the AR model using the PACF.  

- Explain the **Moving Average (MA) model**, and how to identify the order of the MA model using the ACF.  

- Explain the **Autoregressive Moving Average (ARMA) model**.  

- Understand how to fit AR, MA, and ARMA models using the `ARIMA()` method from the [statsforecast](https://nixtlaverse.nixtla.io/statsforecast/docs/models/arima.html) Python package.

---

## 📝 Review and Clarification 📝

1. **Class Notes**: Take some time to revisit your class notes for key insights and concepts.
2. **Zoom Recording**: The recording of today's class will be made available on Canvas approximately 3-4 hours after the session ends.
3. **Questions**: Please don't hesitate to ask for clarification on any topics discussed in class. It's crucial not to let questions accumulate. 

