---
title: "27 - Time Series Regression"
subtitle: "Regression with a Trend and Seasonality"
author: "Fadel M Megahed"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    paged_df: true
    theme: simplex
    code_folding: show
    code_download: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Recap of What we Learned Last Week

  1. Conduct tests for significance of individual regression coefficients.  
  2. Use a simple linear regression model for trend adjustment.
  3. Interpret regression diagnostic plots.  
  4. Create prediction intervals for individual values of the response variable.  
  5. Discuss Exam 03 (Not in R).
 
  
## Required Packages
In the code chunk below, we use the `pacman` package to load the required packages and install them if necessary.

```{r packages}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(tidyverse, 
               magrittr, # for the two-way pipe
               fpp2, # accuracy() function, Arima(), auto.arima(),
               astsa, # package is for the required data
               plotly, # for interactive graphs
               stargazer) 

source('resPlotTS.R') # loading our resPlot function from Last Class
```



## A Refresher of Time-Series Regression Concepts Learned so Far

Using the `jj` dataset from the [astsa package](https://cran.r-project.org/web/packages/astsa/astsa.pdf), we have discussed how we can regress the logged EPS for J&J as a function of time. The regression can be performed using the: (a) `lm()` from base R, where we would need to create the time-feature(s) ourselves, or (b) using the `tslm()` from the forecast package (which we typically load using fpp2). **In class, I will quickly go through the process of fitting the data to ensure that we are all on the same page.**
```{r tsRegRecap}

```

## Main Learning Outcomes for Today's Class
  
  (1) Use regression to account for seasonality in a time series.  
  (2) Compare the performance of the regression model with that obtained from `auto.arima()`.


---


# Seasonal Regression

## Using the `lm()` Function
```{r seasonalReg1}

```


## Using the `tslm()` Function

```{r seasonalReg2}

```


---

# Comparison with auto.arima()

Use the log_jj data.  

  (1)	Use the `auto.arima()` function to fit an arima model to the data.  What model is fit to the data?   
  
  (2) Compare the residuals and the accuracy of the seasonal regression, to the auto.arima results.  Which model fits better?

```{r activity1}


```
