---
title: "ISA 444: Business Forecasting"
subtitle: "04: Plotting a Single Time Series in Python (Cont.)"
author: '<br>Fadel M. Megahed, PhD <br><br> Professor <br> Farmer School of Business<br> Miami University<br><br> [`r icons::icon_style(icons::fontawesome("twitter"), fill = "white")` @FadelMegahed](https://twitter.com/FadelMegahed) <br> [`r icons::icon_style(icons::fontawesome("github"), fill = "white")` fmegahed](https://github.com/fmegahed/) <br> [`r icons::icon_style(icons::fontawesome("paper-plane", style = "solid"), fill = "white")` fmegahed@miamioh.edu](mailto:fmegahed@miamioh.edu)<br> [`r icons::icon_style(icons::fontawesome("question"), fill = "white")` Automated Scheduler for Office Hours](https://calendly.com/fmegahed)<br><br>'
date: "Spring 2025"
output:
  xaringan::moon_reader:
    self_contained: true
    css: [default, "../../style_files/fonts.css", "../../style_files/my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightLanguage: ["python"]
      countIncrementalSlides: false
      ratio: "16:9"
header-includes:  
  - "../../style_files/header.html"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      progress = FALSE, 
                      verbose = FALSE,
                      dev = 'png',
                      fig.height = 2.75,
                      dpi = 300,
                      fig.align = 'center')

options(htmltools.dir.version = FALSE)


miamired = '#C3142D'

if(require(pacman)==FALSE) install.packages("pacman")
if(require(devtools)==FALSE) install.packages("devtools")
if(require(countdown)==FALSE) devtools::install_github("gadenbuie/countdown")
if(require(xaringanExtra)==FALSE) devtools::install_github("gadenbuie/xaringanExtra")
if(require(urbnmapr)==FALSE) devtools::install_github('UrbanInstitute/urbnmapr')
if(require(emo)==FALSE) devtools::install_github("hadley/emo")

pacman::p_load(tidyverse, magrittr, lubridate, janitor, # data analysis pkgs
               DataExplorer, scales, plotly, calendR, pdftools, # plots
               tmap, sf, urbnmapr, tigris, # maps
               bibliometrix, # for bibliometric analysis of my papers
               gifski, av, gganimate, ggtext, glue, extrafont, # for animations
               emojifont, emo, RefManageR, xaringanExtra, countdown) # for slides
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
if(require(xaringanthemer) == FALSE) install.packages("xaringanthemer")
library(xaringanthemer)

style_mono_accent(base_color = "#84d6d3",
                  base_font_size = "20px")

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE  
)

xaringanExtra::use_xaringan_extra(c("tile_view", "animate_css", "tachyons", "panelset", "broadcast", "share_again", "search", "fit_screen", "editable", "clipboard"))

```

## Quick Refresher of Last Class

`r emo::ji("check")` Generate and interpret simple line charts

`r emo::ji("check")` Create seasonal plots and subplots

---

## Learning Objectives for Today's Class

- Describe a lag and create lag scatter plots.  

- Plot and interpret the autocorrelation function (ACF) for a time-series.



---
class: inverse, center, middle

# Describe a lag and create lag scatter plots

---

## What is a Lag?

A **lag** refers to a time delay between an event and its observable effect or measurement.

In time series data, a **lag** is simply a previous observation in the series. For example:  
  + **Lag 1:** Refers to the value immediately preceding the current value.  
  + **Lag 2:** Refers to the value two time periods before the current value, and so on.

---

## What is a Lag (Cont.)?

Imagine a time series of daily sales:  
  + Day 1: $100  
  + Day 2: $150  
  + Day 3: $120  

For Day 3, the lag 1 value is the sales from Day 2 ($150). Similarly, the lag 2 value is from Day 1 ($100). 

In mathematical notation, $y_t$ represents the value at time $t$, then:  
  + $y_{t-1}$ is the lag 1 value of $y_t$  
  + $y_{t-2}$ is the lag 2 value of $y_t$  


---

## Importance of Lags

- **Lags** are important in time series analysis because they can help us understand the relationship between past and present values. They are used in .black[.bold[Forecasting]]:   
     * **Autoregressive models** Models such as AR (AutoRegressive) and ARIMA models use past values (lags) of the series to predict future values. For example, an AR(1) model uses predicts $y_t$ solely based on $y_{t-1}$.  
     * **Seasonality:**  Lags help capture seasonal patterns in data.  

.pull-left[     
```{python example_plot1, echo=FALSE}
import pandas as pd
import matplotlib.pyplot as plt
import matplotlib.patches as mpatches 

df = (
  pd.read_csv('../../data/aus_beer.csv')
  .assign(
    date = lambda x: pd.to_datetime(x['date']),
    quarter = lambda x: x['date'].dt.quarter,
    color = lambda x: x['quarter'].map({1: 'red', 2: 'blue', 3: 'green', 4: 'purple'})
    )
  .query("date >= '2000-01-01'")
  )
    
# Plot the time series
ax = df.plot(x='date', y='beer', legend=False, color = 'gray')
# color by quarter name in scatter in scatter plot
scatter = df.plot(x='date', y='beer', kind='scatter', c=df['color'], ax=ax)

# Create custom legend handles for the quarters
legend_handles = [
    mpatches.Patch(color='red', label='Q1'),
    mpatches.Patch(color='blue', label='Q2'),
    mpatches.Patch(color='green', label='Q3'),
    mpatches.Patch(color='purple', label='Q4')
]

# Add the custom legend to the axes with a title
ax.legend(handles=legend_handles, title="Quarter", loc='upper left')

# Add axis labels and a title
plt.show()

```
]


.pull-right[
```{python acf_out2, echo=FALSE}
from statsmodels.graphics.tsaplots import plot_acf
import matplotlib.pyplot as plt

df = pd.read_csv('../../data/aus_beer.csv')

plot_acf(df['beer'], lags=12)
plt.title('Autocorrelation Function (ACF)')
plt.xlabel('Lag (quarter)')
plt.ylabel('Autocorrelation')
plt.show()
```
]



     

---


## Importance of Lags

- **Lags** are important in time series analysis because they can help us understand the relationship between past and present values. Hence, they are used in:  
  + **Causal Analysis:**  
     * **Cause and Effect:** In many business scenarios, past events (lags) can be predictors for current outcomes; e.g., past advertising spend might affect current sales.  
     * **Policy Evaluation:** Understanding the delayed impact of policies or changes (e.g., price changes, marketing campaigns) is essential.  
     


---

## Importance of Lags (Cont.)

- **Lags** are important in time series analysis because they can help us understand the relationship between past and present values. Hence, they are used in:  
  + **Data Preparation and Feature Engineering:**  
     * **Feature Creation:** In machine learning models, creating lag features from time series data can help the model understand historical patterns and trends.    
     * **Smoothing and Differencing:**  Lags are also used in techniques like differencing to stabilize the mean of a time series, which is a common preprocessing step.  
     

---

## Implementing Lags in Python

In Python, we can create lags using the `shift()` method in pandas.  

```{python lagged_features}
import pandas as pd

df = (
  pd.DataFrame(
    {'date': pd.date_range(start='2025-01-01', periods=5, freq='D'),
    'sales': [100, 150, 120, 200, 180] }
    )
  .assign(
      lag1 = lambda x: x['sales'].shift(1),
      lag2 = lambda x: x['sales'].shift(2)
    )
)

df.head(3)

```
]


---

## Class Activity: Plot the `aus_beer` Time Series

`r countdown(minutes = 5, seconds = 0, top = 0, font_size = "2em")`

+ Load the `aus_beer.csv` dataset from Canvas.  
+ Create a line plot of the `beer` column.  
+ Color the plot by quarter.  
+ Add a legend to the plot.


```{python example_plot1again, echo=FALSE, out.width='70%'}
import pandas as pd
import matplotlib.pyplot as plt
import matplotlib.patches as mpatches 

df = (
  pd.read_csv('../../data/aus_beer.csv')
  .assign(
    date = lambda x: pd.to_datetime(x['date']),
    quarter = lambda x: x['date'].dt.quarter,
    color = lambda x: x['quarter'].map({1: 'red', 2: 'blue', 3: 'green', 4: 'purple'})
    )
  .query("date >= '2000-01-01'")
  )
    
# Plot the time series
ax = df.plot(x='date', y='beer', legend=False, color = 'gray')
# color by quarter name in scatter in scatter plot
scatter = df.plot(x='date', y='beer', kind='scatter', c=df['color'], ax=ax)

# Create custom legend handles for the quarters
legend_handles = [
    mpatches.Patch(color='red', label='Q1'),
    mpatches.Patch(color='blue', label='Q2'),
    mpatches.Patch(color='green', label='Q3'),
    mpatches.Patch(color='purple', label='Q4')
]

# Add the custom legend to the axes with a title
ax.legend(handles=legend_handles, title="Quarter", loc='upper left')

# Add axis labels and a title
plt.show()

```


---


## Demo: Creating and Interpreting Lag Scatter Plots

.pull-left[
In this demo, we will:  
  + Create lag scatter plots for a time series.  
  + Interpret the relationship between the current value and its lagged values.  ]

.pull-right[
```{python lag_scatter_demo, echo=FALSE, out.width ='100%', fig.align='center', fig.keep='last'}
import pandas as pd
import matplotlib.pyplot as plt
import matplotlib.patches as mpatches 
import seaborn as sns

df = (
  pd.read_csv('../../data/aus_beer.csv')
  .assign(
    date = lambda x: pd.to_datetime(x['date']),
    quarter = lambda x: x['date'].dt.quarter,
    color = lambda x: x['quarter'].map({1: 'red', 2: 'blue', 3: 'green', 4: 'purple'})
    )
  .query("date >= '2000-01-01'")
  )

# Create lag columns for lags 1 to 9 using assign
lags = range(1, 9)
for k in lags:
    df = df.assign(**{f'lag_{k}': df['beer'].shift(k)})

lag_cols = [f'lag_{k}' for k in lags]

df_melt = df.melt(
  id_vars=['date', 'beer', 'quarter'],
  value_vars=lag_cols,
  var_name='lag',
  value_name='lag_value'
  )

# Create the facet grid with Seaborn and color-code by month
g = sns.FacetGrid(
  df_melt, col='lag', hue='quarter',
  palette={1: 'red', 2: 'blue', 3: 'green', 4: 'purple'},
  col_wrap=3, height=3, 
  sharex=False, sharey=False
  )
g.map(sns.scatterplot, 'lag_value', 'beer', alpha=0.6 )
# Customize the plot: add axis labels and a main title
g.set_axis_labels("Lagged beer", "beer")
g.fig.suptitle('Lag Plots of beer (Lags 1–8) Colored by Quarter')
g.add_legend(title="Quarter", loc='lower right')

```

]

---
class: inverse, center, middle

# Plot and interpret the ACF for a Time Series

---

## What is the Autocorrelation Function (ACF)?

The **Autocorrelation Function (ACF)** is a measure of the correlation between a time series and a lagged version of itself.

.pull-left[
```{python acf, eval=FALSE}
from statsmodels.graphics.tsaplots import plot_acf
import matplotlib.pyplot as plt

df = pd.read_csv('../../data/aus_beer.csv')

plot_acf(df['beer'], lags=12)
plt.title('Autocorrelation Function (ACF)')
plt.xlabel('Lag (quarter)')
plt.ylabel('Autocorrelation')
plt.show()
```
]


.pull-right[
```{python acf_out, echo=FALSE}
from statsmodels.graphics.tsaplots import plot_acf
import matplotlib.pyplot as plt

# Assume that the data is already loaded in df

plot_acf(df['beer'], lags=12)
plt.title('Autocorrelation Function (ACF)')
plt.xlabel('Lag (quarter)')
plt.ylabel('Autocorrelation')
plt.show()
```
]


---
class: inverse, center, middle

# Recap

---

## Summary of Main Points

By now, you should be able to do the following:  

- Describe a lag and create lag scatter plots.

- Plot and interpret the ACF of a time-series.

---

## 📝 Review and Clarification 📝

1. **Class Notes**: Take some time to revisit your class notes for key insights and concepts.
2. **Zoom Recording**: The recording of today's class will be made available on Canvas approximately 3-4 hours after the session ends.
3. **Questions**: Please don't hesitate to ask for clarification on any topics discussed in class. It's crucial not to let questions accumulate. 
  
